---
title: "Results"
format: html
---

```{r}

#| label: setup-general
#| include: false
#| echo: false
#| warning: false
#| message: false

library(here)
library(gt)
library(dplyr)

source(here::here("helper.R"))

```

## Misunderstanding recruits sustained but not peak effort

```{r}
#| label: setup-torque-cum
#| include: false

# cumulative torque
m3_tc <- readRDS(here::here("../08_Analysis_Modeling/models/h1.m3_torq_cum.rds"))
c1_tc      <- fe(m3_tc, "correction2M1")
c2_tc      <- fe(m3_tc, "correction3M2")
# modality1 raw coefficient (gesture vs grand mean)
mod_tc          <- fe(m3_tc, "modality1")
# gesture vs multimodal full contrast (deviation coding × 2)
mod_ges_mm_tc   <- fe_contrast(m3_tc, "modality1", multiplier = 2)

# instant torque
m3_ti <- readRDS(here::here("../08_Analysis_Modeling/models/h1.m3_torq_in.rds"))
re_ppt_mod_ti <- re_sd(m3_ti, "pcn_ID", "modality1")

# cumulative envelope
m3_ec <- readRDS(here::here("../08_Analysis_Modeling/models/h1.m3_env_cum.rds"))
c1_ec      <- fe(m3_ec, "correction2M1")
c2_ec      <- fe(m3_ec, "correction3M2")
mod_ec     <- fe_contrast(m3_ec, "modality1", multiplier = 2)

# instant envelope
m3_ei <- readRDS(here::here("../08_Analysis_Modeling/models/h1.m3_env_in.rds"))
mod_ei     <- fe_contrast(m3_ei, "modality1", multiplier = 2)

# cumultive COPc 
m3_cc <- readRDS(here::here("../08_Analysis_Modeling/models/h1.m3_cop_cum.rds"))
c1_cc      <- fe(m3_cc, "correction2M1")
c2_cc      <- fe(m3_cc, "correction3M2")
go_vs_vo_cc <- fe_derived_contrast(m3_cc, "modality1", "modality2", w1 = 2,  w2 = 1)
mm_vs_vo_cc <- fe_derived_contrast(m3_cc, "modality1", "modality2", w1 = 1,  w2 = 2)

# instant COPc 
m3_ci <- readRDS(here::here("../08_Analysis_Modeling/models/h1.m3_cop_in.rds"))
go_vs_vo_ci <- fe_derived_contrast(m3_ci, "modality1", "modality2", w1 = 2,  w2 = 1)
mm_vs_vo_ci <- fe_derived_contrast(m3_ci, "modality1", "modality2", w1 = 1,  w2 = 2)
mm_vs_go_ci <- fe_derived_contrast(m3_ci, "modality1", "modality2", w1 = -1, w2 = 1)

# torque and time
e2_m <- readRDS(here::here("../08_Analysis_Modeling/models/e2.m1_torq_cum.rds"))
c1_tc_wt      <- fe(e2_m, "correction2M1")

# first trial effort
e2_in_torq  <- readRDS(here::here("../08_Analysis_Modeling/models/e1.m1_torq_in.rds"))
e2_in_env   <- readRDS(here::here("../08_Analysis_Modeling/models/e1.m1_env_in.rds"))
e2_in_cop   <- readRDS(here::here("../08_Analysis_Modeling/models/e1.m1_cop_in.rds"))

# ── Correction main effects ────────────────────────────────────────────────────
c1_in_torq  <- fe(e2_in_torq,  "correction2M1")
c1_in_env   <- fe(e2_in_env,   "correction2M1")
c1_in_cop   <- fe(e2_in_cop,   "correction2M1")

# ── Interaction at ±2SD of baseline ───────────────────────────────────────────
# Check exact param names with: rownames(fixef(e2_in_torq))
# Adjust interaction_param strings below to match

# Instantaneous torque × baseline
int_in_torq_lo <- fe_at_baseline(e2_in_torq, "correction2M1",
                                  "correction2M1:baseline_torque_in_z", -2)
int_in_torq_hi <- fe_at_baseline(e2_in_torq, "correction2M1",
                                  "correction2M1:baseline_torque_in_z",  2)

# Instantaneous envelope × baseline
int_in_env_lo  <- fe_at_baseline(e2_in_env, "correction2M1",
                                  "correction2M1:baseline_env_in_z",  -2)
int_in_env_hi  <- fe_at_baseline(e2_in_env, "correction2M1",
                                  "correction2M1:baseline_env_in_z",   2)

# Instantaneous COP × baseline
int_in_cop_lo  <- fe_at_baseline(e2_in_cop, "correction2M1",
                                  "correction2M1:baseline_cop_in_z",  -2)
int_in_cop_hi  <- fe_at_baseline(e2_in_cop, "correction2M1",
                                  "correction2M1:baseline_cop_in_z",   2)

```

As predicted, participants invested more effort when their partners misunderstood them – but the pattern was more specific than a simple mobilization of all available resources. Across all investigated dimensions of effort – upper-limb, vocal, and postural – behavior following an incorrect guess was more effortful in cumulative but not instantaneous terms (Fix X). Upper-limb effort showed the largest enhancement: cumulative torque change increased by `r pct(c1_tc$pct)` at the first correction (95% CrI: `r cri_pct(c1_tc$pct_lo, c1_tc$pct_hi)`) and by `r pct(c2_tc$pct)` at the second correction (95% CrI: `r cri_pct(c2_tc$pct_lo, c2_tc$pct_hi)`). Vocal and postural effort followed the same pattern at roughly half the magnitude. Cumulative vocal effort increase by `r pct(c1_ec$pct)` at the first correction (95% CrI: `r cri_pct(c1_ec$pct_lo, c1_ec$pct_hi)`) and by `r pct(c2_ec$pct)` at the second (95% CrI: `r cri_pct(c2_ec$pct_lo, c2_ec$pct_hi)`), and cumulative postural effort increased by `r pct(c1_cc$pct)` at the first correction (95% CrI: `r cri_pct(c1_cc$pct_lo, c1_cc$pct_hi)`) and by `r pct(c2_cc$pct)` at the second (95% CrI: `r cri_pct(c2_cc$pct_lo, c2_cc$pct_hi)`). 

To test whether these cumulative increases reflected more movement or larger movements, we refitted the models with trial duration as a covariate. Adding duration substantially attenuated the correction effect (cf. Supplementary Material). The first-correction effect on the upper-limb effort remained credibly positive, but decreased from `r pct(c1_tc$pct)` to `r pct(c1_tc_wt$pct)` (95% CrI: `r cri_pct(c1_tc_wt$pct_lo, c1_tc_wt$pct_hi)`), suggesting that almost two-thirds of the increase in upper-limb effort can be attributed to accumulation of more, rather than larger, arm movement. The second-correction effect  on upper-limb effort and both corrections effects for vocal and postural effort, were fully accounted for by duration: with duration in the model, none remained credibly different from zero. 

None of the instantaneous measures showed a similar credible pattern: at the populational level, the peak vocal, upper-limb, and postural effort remained unchanged following a misunderstanding. In an exploratory analysis, however, a model including the interaction between correction and each performer’s average baseline (i.e., first performance) effort showed that the first-correction effect is credibly influenced by the amount of effort in the baseline performance, and can even lead in the opposite direction (see Supplementary Material section X). Specifically, for peak torque change, the interaction with baseline effort was strongly negative: performers with low average baseline effort (−2 SD from the mean) increased their  first correction by `r fmt_fe_pct(int_in_torq_lo)`, while those with high baseline effort (+2 SD) decreased it by `r fmt_fe_pct(int_in_torq_hi)`. Instantaneous vocal effort showed the same dependance: low-baseline performers (-2 SD) increased their first correction by `r fmt_fe_pct(int_in_env_lo)`, while high-baseline performers (+2 SD) decreased it by `r fmt_fe_pct(int_in_env_hi)`. For postural effort, the pattern was weaker: high-baseline performers decreased their first correction by `r fmt_fe_pct(int_in_cop_hi)` in the first correction, while low-baseline performers showed no credible change (`r fmt_fe_pct(int_in_cop_lo)`). Instantaneous effort at correction was therefore not uniformly recruited; whether performers scaled up the peak effort depended on how effortful their baseline performance already was. We return to the interpretation of this baseline dependence in the Discussion.

We also asked how baseline effort levels differed across modality conditions, since the available articulators differ between them. Modality shaped effort, but in measure-specific ways. Vocal effort was substantially higher in the vocal-only than in the multimodal condition, both cumulatively (`r pct(mod_ec$pct)` higher, 95% CrI `r cri_pct(mod_ec$pct_lo, mod_ec$pct_hi)`) and instantaneously (`r pct(mod_ei$pct)` higher, 95% CrI `r cri_pct(mod_ei$pct_lo, mod_ei$pct_hi)`), indicating that vocal effort takes on more of the communicative burden when the hands are unavailable. Postural-effort modality effect is reversed in direction: conditions involving bodily movement elicited far greater COPc than vocal-only, both cumulatively (gesture-only: `r pct(go_vs_vo_cc$pct)` higher, 95% CrI: `r cri_pct(go_vs_vo_cc$pct_lo, go_vs_vo_cc$pct_hi)`; multimodal: `r pct(mm_vs_vo_cc$pct)` higher, 95% CrI: `r cri_pct(mm_vs_vo_cc$pct_lo, mm_vs_vo_cc$pct_hi)`) and instantaneously (gesture-only: `r pct(go_vs_vo_ci$pct)` higher, 95% CrI: `r cri_pct(go_vs_vo_ci$pct_lo, go_vs_vo_ci$pct_hi)`, multimodal: `r pct(mm_vs_vo_ci$pct)` higher,
95% CrI: `r cri_pct(mm_vs_vo_ci$pct_lo, mm_vs_vo_ci$pct_hi)`). Gesture-only and multimodal conditions did not differ in cumulative postural effort, although multimodal exceeded gesture-only for peak postural effort, suggesting that coordinating two modalities recruits bursts of postural activity without increasing total postural load. For upper-limb effort, we found no reliable average difference between conditions once between-participant and between-concept variability was modelled.The gesture-only vs. multimodal contrast was positive for cumulative effort but its interval included zero (`r pct(mod_tc$pct)`, 95% CRI `r cri_pct(mod_tc$pct_lo, mod_tc$pct_hi)`). Substantial random slope variance indicated that who is communicating and which concept is conveyed matter more for upper-limb effort than the modality condition itself.

To summarize, while cumulative effort increased reliably across corrections for all three measures, instantaneous effort did not. Correction therefore imposes a clear cumulative cost on performers without recruiting greater peak intensity per movement or per vocalization. This pattern is consistent with a redistribution of effort across longer time intervals and more movement units rather than a concentrated scaling-up of force or amplitude, an interpretation we develop in the Discussion. The consistency of these increases across three independent measures of bodily effort indicates that repair imposes a substantive physical cost on the performer, supporting H1 for cumulative, though not instantaneous, effort.

## Upper-limb effort scales with the degree of misunderstanding

```{r}
#| label: setup-prev-answer
#| include: false


h2m_ct <- readRDS(here::here("../08_Analysis_Modeling/models/h2.m1b_torq_cum.rds"))
h2m_it <- readRDS(here::here("../08_Analysis_Modeling/models/h2.m1b_torq_in.rds"))

prev_an_ct_ci <- fe(h2m_ct, "answer_prev_dist_z")
prev_an_it_ci <- fe(h2m_it, "answer_prev_dist_z")

```

We found partial support for H2. Cumulative upper-limb effort scaled credibly with the degree of misunderstanding, operationalized as the cosine similarity between the guesser’s previous answer and the target meaning (lower similarity indicates greater misunderstanding). Each 1-SD increase in cosine similarity was associated with `r pct(prev_an_ct_ci$pct)` reduction in cumulative torque change (95% CrI: `r cri_pct(prev_an_ct_ci$pct_lo, prev_an_ct_ci$pct_hi)`); equivalently, performers invested more cumulative upper-limb effort when the previous answer was further from the target. Instantaneous torque change ran in the same direction, but its estimated interval included zero (`r fmt_fe_pct(prev_an_it_ci)`). In contrast, neither cumulative nor instantaneous vocal effort showed a credible relationship with cosine similarity, and the same was true for both postural measures.

Cumulative upper-limb effort therefore adapts reliably to the amount of repair work the previous answer requires, while vocal and postural effort appear insensitive to the degree of misunderstanding. The asymmetry between sustained and peak upper-limb effort mirrors the pattern reported in the previous subsection: misunderstanding scales cumulative effort more reliably than peak effort.

Importantly, the upper-limb cosine similarity effect was uniformly distributed across the sample: the random-slope variance for cosine similarity was negligible across all three grouping levels (concepts, participants, and dyads), indicating that the effect did not vary systematically by performer, dyad, or concept. This consistency suggests that arm movement responds to common ground in a relatively automatic way, rather than reflecting strategic adaptation by particular individuals or for particular concepts.

## Effort does not predict communicative success


```{r extract-all, include=FALSE}

# ── Models ────────────────────────────────────────────────────────────────────
e3_m1        <- readRDS(here::here("../08_Analysis_Modeling/models/e3.m1.rds"))
e3_m1b       <- readRDS(here::here("../08_Analysis_Modeling/models/e3.m1b.rds"))
fit_sim      <- readRDS("../08_Analysis_Modeling/models/e3.m3_att.rds")
fit_sim_inst <- readRDS("../08_Analysis_Modeling/models/e3.m3b_att.rds")
fit_ord      <- readRDS("../08_Analysis_Modeling/models/e3.m4_att.rds")
fit_ord_inst <- readRDS("../08_Analysis_Modeling/models/e3.m4b_att.rds")
fit_ever     <- readRDS("../08_Analysis_Modeling/models/e3.m7_att.rds")
fit_ever_inst <- readRDS("../08_Analysis_Modeling/models/e3.m7b_att.rds")
fit_m8       <- readRDS("../08_Analysis_Modeling/models/e3.m8_att.rds")
fit_m8b      <- readRDS("../08_Analysis_Modeling/models/e3.m8b_att.rds")

CAT_C0    <- 1
CAT_NEVER <- 4

# ── First-attempt resolution (cumulative) ─────────────────────────────────────
nd_base      <- make_nd_base(e3_m1)
epred_base   <- posterior_epred(e3_m1, newdata = nd_base, re_formula = NA)
idx          <- setNames(seq_along(nd_base$modality), nd_base$modality)

p_gest_base  <- ci(epred_base[, idx["gesture"]])
p_mm_base    <- ci(epred_base[, idx["multimodal"]])
p_voc_base   <- ci(epred_base[, idx["vocal"]])

dp_gest_vs_mm <- ci(epred_base[, idx["gesture"]]    - epred_base[, idx["multimodal"]])
dp_voc_vs_mm  <- ci(epred_base[, idx["vocal"]]      - epred_base[, idx["multimodal"]])

# ── Answer similarity (cumulative) ────────────────────────────────────────────
nd_base_sim <- make_nd_base(
  fit_sim,
  extra_zero_vars   = "answer_prev_dist_z",
  extra_factor_refs = list(correction = levels(fit_sim$data$correction)[1])
)
epred_base_sim <- posterior_epred(fit_sim, newdata = nd_base_sim, re_formula = NA)
idx_sim        <- setNames(seq_along(nd_base_sim$modality), nd_base_sim$modality)

p_gest_sim_base <- ci(epred_base_sim[, idx_sim["gesture"]])
p_mm_sim_base   <- ci(epred_base_sim[, idx_sim["multimodal"]])
p_voc_sim_base  <- ci(epred_base_sim[, idx_sim["vocal"]])

dp_gest_vs_mm_sim <- ci(epred_base_sim[, idx_sim["gesture"]]    - epred_base_sim[, idx_sim["multimodal"]])
dp_voc_vs_mm_sim  <- ci(epred_base_sim[, idx_sim["vocal"]]      - epred_base_sim[, idx_sim["multimodal"]])

# ── Resolution speed / ordinal (cumulative) ───────────────────────────────────
nd_base_ord <- make_nd_base(fit_ord)
epred_ord   <- posterior_epred(fit_ord, newdata = nd_base_ord, re_formula = NA)
idx_ord     <- setNames(seq_along(nd_base_ord$modality), nd_base_ord$modality)

pc0_gest_ord  <- ci(ord_cat(epred_ord, idx_ord["gesture"],    CAT_C0))
pc0_mm_ord    <- ci(ord_cat(epred_ord, idx_ord["multimodal"], CAT_C0))
pc0_voc_ord   <- ci(ord_cat(epred_ord, idx_ord["vocal"],      CAT_C0))
pnev_mm_ord   <- ci(ord_cat(epred_ord, idx_ord["multimodal"], CAT_NEVER))
pnev_voc_ord  <- ci(ord_cat(epred_ord, idx_ord["vocal"],      CAT_NEVER))

dp_c0_gest_vs_mm_ord <- ci(ord_cat(epred_ord, idx_ord["gesture"],    CAT_C0) -
                            ord_cat(epred_ord, idx_ord["multimodal"], CAT_C0))
dp_c0_voc_vs_mm_ord  <- ci(ord_cat(epred_ord, idx_ord["vocal"],      CAT_C0) -
                            ord_cat(epred_ord, idx_ord["multimodal"], CAT_C0))

# Credible interaction: envelope × vocal (instantaneous ordinal)
fe_ord_inst          <- fixef(fit_ord_inst, summary = FALSE)
nd_base_ord_inst     <- make_nd_base_in(fit_ord_inst)
epred_ord_inst       <- posterior_epred(fit_ord_inst, newdata = nd_base_ord_inst, re_formula = NA)

b_env_x_voc_ord_inst      <- ci(fe_ord_inst[, "modality3:envelope_in_log_c"])
prod_c0_env_voc_ord_inst  <- productivity_ord(fit_ord_inst, nd_base_ord_inst, "vocal", "envelope_in_log_c", CAT_C0)
prod_nev_env_voc_ord_inst <- productivity_ord(fit_ord_inst, nd_base_ord_inst, "vocal", "envelope_in_log_c", CAT_NEVER)

# ── Ever resolved ─────────────────────────────────────────────────────────────
nd_base_ever <- make_nd_base(fit_ever)
epred_ever   <- posterior_epred(fit_ever, newdata = nd_base_ever, re_formula = NA)
idx_ever     <- setNames(seq_along(nd_base_ever$modality), nd_base_ever$modality)

p_gest_ever <- ci(epred_ever[, idx_ever["gesture"]])
p_mm_ever   <- ci(epred_ever[, idx_ever["multimodal"]])
p_voc_ever  <- ci(epred_ever[, idx_ever["vocal"]])

dp_gest_vs_mm_ever <- ci(epred_ever[, idx_ever["gesture"]]    - epred_ever[, idx_ever["multimodal"]])
dp_voc_vs_mm_ever  <- ci(epred_ever[, idx_ever["vocal"]]      - epred_ever[, idx_ever["multimodal"]])

# Credible interactions (instantaneous ever-resolved)
nd_base_ever_inst  <- make_nd_base_in(fit_ever_inst)
epred_ever_inst    <- posterior_epred(fit_ever_inst, newdata = nd_base_ever_inst, re_formula = NA)
idx_ever_inst      <- setNames(seq_along(nd_base_ever_inst$modality), nd_base_ever_inst$modality)

fe_ever_inst           <- fixef(fit_ever_inst, summary = FALSE)
b_env_x_mm_ever_inst   <- ci(fe_ever_inst[, "modality1:envelope_in_log_c"])
b_copc_x_voc_ever_inst <- ci(fe_ever_inst[, "modality3:copc_in_log_c"])

prod_ever_inst_env_mm   <- productivity(fit_ever_inst, nd_base_ever_inst, "multimodal", "envelope_in_log_c")
prod_ever_inst_copc_voc <- productivity(fit_ever_inst, nd_base_ever_inst, "vocal",      "copc_in_log_c")

p_voc_ever_inst_copc_lo <- ci(pred_at(fit_ever_inst, nd_base_ever_inst, "vocal", "copc_in_log_c", -1))
p_voc_ever_inst_copc_hi <- ci(pred_at(fit_ever_inst, nd_base_ever_inst, "vocal", "copc_in_log_c",  1))

# ── Total effort / final similarity (m8 cumulative) ───────────────────────────
nd_base_m8 <- fit_m8$data |>
  group_by(modality) |> slice(1) |> ungroup() |>
  mutate(
    arm_torque_total_c = 0, envelope_total_c = 0, copc_total_c = 0,
    concept_number_c = 0, answer_prev_dist_z = 0, expressibility_z = 0,
    Familiarity = mean(fit_m8$data$Familiarity),
    BFI_extra   = mean(fit_m8$data$BFI_extra)
  )

epred_m8 <- posterior_epred(fit_m8, newdata = nd_base_m8, re_formula = NA)
idx_m8   <- setNames(seq_along(nd_base_m8$modality), nd_base_m8$modality)

dp_gest_vs_mm_m8 <- ci(epred_m8[, idx_m8["gesture"]]    - epred_m8[, idx_m8["multimodal"]])
dp_voc_vs_mm_m8  <- ci(epred_m8[, idx_m8["vocal"]]      - epred_m8[, idx_m8["multimodal"]])
```

In an exploratory analysis, we asked whether the effort performers invested in repair translated into improved communicative outcomes. Across four complementary outcomes (1. whether the misunderstanding was resolved on the first attempt, 2. how semantically similar the guesser’s answer was to the target concept, 3. how quickly resolution was reached across correction attempts, and 4. whether resolution was ever achieved across the full sequence), modality was the dominant predictor and effort was not. 

Multimodal trials outperformed vocal-only trials across all outcomes. The predicted probability of first-attempt resolution was `r fmt_p(p_mm_base["med"], p_mm_base["lo"], p_mm_base["hi"])` for multimodal
and `r fmt_p(p_voc_base["med"], p_voc_base["lo"], p_voc_base["hi"])` for vocal-only (contrast = `r fmt_dp(dp_voc_vs_mm["med"])`, 95% CrI [`r fmt_dp(dp_voc_vs_mm["lo"])`, `r fmt_dp(dp_voc_vs_mm["hi"])`]), and the probability of never resolving was (P(never) = `r fmt_p(pnev_mm_ord["med"], pnev_mm_ord["lo"], pnev_mm_ord["hi"])`
for multimodal and `r fmt_p(pnev_voc_ord["med"], pnev_voc_ord["lo"], pnev_voc_ord["hi"])` for vocal-only). Gesture-only performance was largely indistinguishable from multimodal, with the one exception: gesture-only resolved misunderstanding more slowly than multimodal (contrast = `r fmt_dp(dp_c0_gest_vs_mm_ord["med"])`, 95% CrI [`r fmt_dp(dp_c0_gest_vs_mm_ord["lo"])`, `r fmt_dp(dp_c0_gest_vs_mm_ord["hi"])`]). On all other outcomes, gesture-only versus multimodal contrasts remained non-credible. 

Despite the clear effort increase during repair, this investment does not translate into communicative gain. Effort slopes were small, inconsistent in direction, and non-credible across the large majority of effort × modality combinations. A small number of exceptions ran in opposing directions and warrant cautious interpretation. Instantaneous vocal amplitude predicted faster resolution in vocal-only trials, and greater postural displacement at first attempt predicted higher probability of eventual resolution in vocal-only trials – possibly indexing speaker commitment when gesture is unavailable. In the opposite direction, higher peak vocal amplitude in multimodal trials was associated with lower probability of eventual resolution, consistent with louder vocalization marking difficulty rather than driving success.

# Discussion

In this study, we identified an effort-success paradox in communicative repair: when faced with misunderstanding, performers sustain more effort across all three channels of the body, yet this additional investment does not improve communicative outcomes. Behind the paradox lies a specific finding about effort itself. When people encounter misunderstanding in resolving a referent using non-linguistic means, they do not simply do more of the same; they do more over time. This distinction, between elaboration and intensification, may reflect a general principle of how the body responds to communicative demand: not by concentrating effort, but by distributing it across a longer temporal window. Our findings instantiate this principle in a controlled communicative setting, but its implications extend beyond repair. Wherever a communicative attempt fails to achieve its intended effect – whether through noise, inattention, or epistemic misalignment – the performer faces a choice between doing the same thing harder or doing more of it. Our data suggest that the latter is the case for misunderstanding in a game-like referential task, and that it is flexible: sensitive to how large the misunderstanding is, and shaped by what articulatory resources are available.

These findings extend existing accounts of how environmental constraints reshape communicative behavior. Lindblom’s H\&H Theory [@lindblomExplainingPhoneticVariation1990] predicts hyper-articulation under a constraint: a system under pressure produces a clearer, more precise signal. Rather than contradicting this prediction, our results reveal that misunderstanding imposes a categorically different kind of constraint. Situations with background noise demand a stronger signal and local enhancement (e.g., louder speech in a noisy room). Epistemic misalignment demands enhancement that is more global: adding more information and changing the structure of the message. This is in line with existing work on repair [@rasenbergMultimodalNatureCommunicative2022; @hilliardBridgingGapsCommon2016] but our operationalization allows us to claim – for the first time – that performers invest more physical effort when their addressee signals poor understanding, rather than simply reorganizing the content of the message.

Crucially, elaboration is not reflected in the same manner across the three dimensions. Upper-limb movement shows the greatest capacity for sustained increase, as gestures offer multiple routes through which more structure can be added including pointing, tracing, or representing. Vocal effort, on the other hand, showed a more modest increase, possibly reflecting the limited degrees of freedom available for elaborating non-linguistic vocalization. Vocal signals are expressive, but once a referential breakdown has occurred there is no equivalent to pointing or distinguishing an object from an action; the semantic space is harder to navigate without articulatory tools that map onto referents. Postural effort  also showed a cumulative increase across modality conditions, indicating whole-body engagement; the smaller size of this increase compared with the upper-limb effect is consistent with a more basic, stability-oriented functions [@massionMovementPostureEquilibrium1992]. That the postural adjustment pattern appeared even in the vocal-only condition, where communicative arm movements are absent, suggests that misunderstanding imposes demands on the whole body, possibly also reflecting cognitive effort or self-regulatory mechanisms [@yardleyEffectArticulatoryMental1999; @linStandTallThink2026].

The diverging articulatory and meaning-making affordances in gestural and vocal modality is exactly what should make them work best in combination, both in terms of saving effort and reaching success. Our data partially supports that. Vocal effort was substantially lower in the multimodal than in the vocal-only condition, indicating that the voice carries less of the communicative load when the hands are available. The reverse comparison ran in the same direction (smaller upper-limb effort when voice is present) but its interval included zero, suggesting that the arms benefit the voice more than the voice benefits the arms. On the success side, gesture and vocalization together outperformed vocalization alone across all four outcomes. Compared with gesture-only, multimodality offered a credible advantage only on resolution speed; first-attempt resolution, semantic similarity, and eventual resolution were comparable. The multimodal benefit on resolution speed echoes  findings of Macuch Silva et al. [@macuchsilvaMultimodalityOriginNovel2020], corroborating the crucial role of multimodality in efficient (i.e., faster) repair; the absence of a credible advantage on whether resolution is reached suggests that combining channels accelerates repair without necessarily changing its eventual outcome.

The effort-success paradox is clearest in this disconnect: although performers recruited more effort under misunderstanding, that effort did not lead to better outcomes. Literature partially helps to navigate this puzzle. In a recent review, Wolff and colleagues [@wolffWhenEffortFails2025a] noted that while effort–performance coupling is a phenomenon observed across many domains, there are conditions under which effort does not translate into success. The absence of coupling in our data might be hinting at a distinct role of effort in repair. In a noisy room, a stronger signal is a precondition for any interaction to take place at all; misunderstanding may call for something else. Instead of more sustained effort, the interlocutor could benefit from change of tempo, greater informativity of individual units, or – as our data show – the number of communicative channels available. The persistent effect of expressibility on both effort and success in our data points in the same direction: semantic content shapes how effortful a specific meaning is to communicate and how readily it is understood, somewhat independent of how hard the performer works.

In fact, researchers studying noise or misunderstanding also report diverse (if any) patterns: while some hint at more assumed effort bringing more success [@drijversVisualContextEnhanced2017], many leave the question of the actual success unresolved [@rasenbergMultimodalNatureCommunicative2022; @trujilloSpeakersExhibitMultimodal2021; @hollerExperimentalInvestigationHow2011] or fail to find this relationship [@bogelsTherePowerLaw2024]. This question remains open for targeted investigation, but it does seem that the communicative value of multimodality lies not in effort efficiency, but in the richer articulatory palette it affords. 

Yet this does not mean that effort is decoupled from the communicative context altogether. In our second pre-registered analysis, the relationship runs in the other direction: rather than effort driving outcomes, effort responds to them. Cumulative upper-limb effort scaled with the degree of misunderstanding, decreasing as the guesser’s previous answer approached the target meaning. This calibration was specific to the arms; neither vocal nor postural effort tracked the degree of misunderstanding. It was also uniform across the sample, varying little across performers, dyads, or concepts, which suggests that the adjustment of arm effort to common ground operates similarly across individuals rather than reflecting idiosyncratic strategy. Of the three channels, the arms thus emerge as the one whose sustained effort is finely tuned to how far the interaction stands from mutual understanding. This sensitivity, like the correction effect in H1, was specific to sustained effort: peak measures were largely unmoved by misunderstanding in both analyses. The consistent picture is that misunderstanding reshapes how long and how much performers communicate, not how forcefully.

Effort thus responds to communicative feedback rather than driving communicative outcomes – a distinction that may reflect the difference between effort as a reactive, demand-driven phenomenon and effort as a purposive strategy. Together, the absence of an effort-success link and the responsiveness of effort to feedback point toward a reframing of what communicative effort represents. Rather than a resource allocated toward resolution, effort appears to function as a continuous index of the interaction’s epistemic state – signaling mobilization of bodily resources when interaction breaks down, scaling with the distance from common ground, recalibrating when that distance closes, but not itself closing it.

Our findings must be interpreted in light of several limitations. The controlled design of our paradigm, while necessary for isolating effort modulation, comes with trade-offs in ecological validity. Whole-body movement without speech amplifies physical demands beyond what typical face-to-face interaction requires, and in naturalistic settings linguistic structure carries much of the repair work. We view this as a necessary first step: charade-like referential tasks share fundamental properties with linguistic interaction – they are inherently communicative, meaningful, and intersubjective [@christiansenLanguageGameHow2022] – and identifying effort modulation in this simplified context establishes a foundation for investigating these phenomena in more naturalistic settings. 

A related limitation concerns what counts as communicative success. Our design treats resolution as an exact match between the guesser's answer and the target concept, whereas real interaction typically requires only sufficient mutual understanding to support the next action. Future work might instead orient toward more instrumental success criteria – understanding that someone wants a drink, rather than identifying the precise concept of drinking, glass, or water. Our measure of success is also potentially affected by the asymmetric distribution of communicative responsibility in this task: the guesser contributes only a single guess per performance, leaving most of the repair work to the performer. Research on referential tasks with freer interaction suggests that more distributed effort – where partners jointly signal and respond to misunderstanding – may be what drives resolution [@rasenbergMultimodalNatureCommunicative2022]. It is therefore possible that effort does not help in our paradigm precisely because the partner who could benefit from it has limited room to act on it. 

A second set of limitations concerns the measurement of effort itself. The three features we selected – arm torque, acoustic amplitude envelope, and centre of pressure displacement – capture general biomechanical demands on the body, but effort is content-dependent: concepts vary substantially in how much and what kind of effort their communication requires, as reflected in the large concept-level variance in our models. For some meanings, repair may call not for more intensity but for less – communicating ‘silence’ or ‘stillness’ being an obvious case. While our population-level effects suggest that the selected features capture a general mechanism, future work should examine whether other dimensions of movement – speed, spatial precision, or inter-articulator coordination – carry additional information about effort that our measures do not. In particular, the coordinative cost of orchestrating multiple articulators simultaneously remains unaddressed here and may represent a distinct dimension of multimodal effort with its own scaling properties. 

Taken together, these limitations point toward productive directions for future research without undermining the core finding: that the body responds to misunderstanding with sustained, distributed effort – reliably, across modalities, and, for the arms, in proportion to the degree of communicative failure. This effort-success paradox sits in tension with accounts of effort as a minimized resource. If people generally avoid effortful behavior and experience frustration when tasks become demanding, why would they increase physical investment in response to misunderstanding – without clear returns in referent resolution? One possibility is that effort in social interaction is not primarily optimized toward task success but toward something else: signaling engagement, maintaining the interaction, or affirming commitment to the partner. Research on challenging tasks suggests that effort itself can carry social value, increasing the perceived worth of its product including social belonging[@inzlichtEffortParadoxEffort2018]. On this view, the effort enhancement we observe may be less about solving the problem and more about staying in it together.
