---
title: "Results"
format: html
execute:
  enabled: true
---

```{r}

#| label: setup-general
#| include: false
#| echo: false
#| warning: false
#| message: false

library(here)
library(gt)
library(dplyr)

source(here::here("Manuscript/helper.R"))

```

## Misunderstanding recruits cumulative but not instantaneous effort

```{r}
#| label: setup-torque-cum
#| include: false

# cumulative torque
m3_tc <- readRDS(here::here("06_Analysis_Modeling/models/h1.m3_torq_cum.rds"))
c1_tc      <- fe(m3_tc, "correction2M1")
c2_tc      <- fe(m3_tc, "correction3M2")
# modality1 raw coefficient (gesture vs grand mean)
mod_tc          <- fe(m3_tc, "modality1")
# gesture vs multimodal full contrast (deviation coding × 2)
mod_ges_mm_tc   <- fe_contrast(m3_tc, "modality1", multiplier = 2)

# instant torque
m3_ti <- readRDS(here::here("06_Analysis_Modeling/models/h1.m3_torq_in.rds"))
re_ppt_mod_ti <- re_sd(m3_ti, "pcn_ID", "modality1")

# cumulative envelope
m3_ec <- readRDS(here::here("06_Analysis_Modeling/models/h1.m3_env_cum.rds"))
c1_ec      <- fe(m3_ec, "correction2M1")
c2_ec      <- fe(m3_ec, "correction3M2")
mod_ec     <- fe_contrast(m3_ec, "modality1", multiplier = 2)

# instant envelope
m3_ei <- readRDS(here::here("06_Analysis_Modeling/models/h1.m3_env_in.rds"))
mod_ei     <- fe_contrast(m3_ei, "modality1", multiplier = 2)

# cumultive COPc 
m3_cc <- readRDS(here::here("06_Analysis_Modeling/models/h1.m3_cop_cum.rds"))
c1_cc      <- fe(m3_cc, "correction2M1")
c2_cc      <- fe(m3_cc, "correction3M2")
go_vs_vo_cc <- fe_derived_contrast(m3_cc, "modality1", "modality2", w1 = 2,  w2 = 1)
mm_vs_vo_cc <- fe_derived_contrast(m3_cc, "modality1", "modality2", w1 = 1,  w2 = 2)

# instant COPc 
m3_ci <- readRDS(here::here("06_Analysis_Modeling/models/h1.m3_cop_in.rds"))
go_vs_vo_ci <- fe_derived_contrast(m3_ci, "modality1", "modality2", w1 = 2,  w2 = 1)
mm_vs_vo_ci <- fe_derived_contrast(m3_ci, "modality1", "modality2", w1 = 1,  w2 = 2)
mm_vs_go_ci <- fe_derived_contrast(m3_ci, "modality1", "modality2", w1 = -1, w2 = 1)

# torque and time
e2_m <- readRDS(here::here("06_Analysis_Modeling/models/e2.m1_torq_cum.rds"))
c1_tc_wt      <- fe(e2_m, "correction2M1")

# first trial effort
e2_in_torq  <- readRDS(here::here("06_Analysis_Modeling/models/e1.m1_torq_in.rds"))
e2_in_env   <- readRDS(here::here("06_Analysis_Modeling/models/e1.m1_env_in.rds"))
e2_in_cop   <- readRDS(here::here("06_Analysis_Modeling/models/e1.m1_cop_in.rds"))

# ── Correction main effects ────────────────────────────────────────────────────
c1_in_torq  <- fe(e2_in_torq,  "correction2M1")
c1_in_env   <- fe(e2_in_env,   "correction2M1")
c1_in_cop   <- fe(e2_in_cop,   "correction2M1")

# ── Interaction at ±2SD of baseline ───────────────────────────────────────────
# Check exact param names with: rownames(fixef(e2_in_torq))
# Adjust interaction_param strings below to match

# Instantaneous torque × baseline
int_in_torq_lo <- fe_at_baseline(e2_in_torq, "correction2M1",
                                  "correction2M1:baseline_torque_in_z", -2)
int_in_torq_hi <- fe_at_baseline(e2_in_torq, "correction2M1",
                                  "correction2M1:baseline_torque_in_z",  2)

# Instantaneous envelope × baseline
int_in_env_lo  <- fe_at_baseline(e2_in_env, "correction2M1",
                                  "correction2M1:baseline_env_in_z",  -2)
int_in_env_hi  <- fe_at_baseline(e2_in_env, "correction2M1",
                                  "correction2M1:baseline_env_in_z",   2)

# Instantaneous COP × baseline
int_in_cop_lo  <- fe_at_baseline(e2_in_cop, "correction2M1",
                                  "correction2M1:baseline_cop_in_z",  -2)
int_in_cop_hi  <- fe_at_baseline(e2_in_cop, "correction2M1",
                                  "correction2M1:baseline_cop_in_z",   2)

```

As predicted, participants invested more effort when their partners misunderstood them but the pattern was more specific than a simple mobilization of all available resources. Here we report the main effects of interest, a full report of the results is available in Supplementary Material (section F). Across all investigated dimensions of effort – upper-limb, vocal, and postural – behavior following an incorrect guess was more effortful in cumulative but not instantaneous terms (Fig. 1b). Upper-limb effort showed the largest enhancement: cumulative torque change increased by `r pct(c1_tc$pct)` at the first correction compared to baseline (95% CrI: `r cri_pct(c1_tc$pct_lo, c1_tc$pct_hi)`) and by `r pct(c2_tc$pct)` at the second correction compared to first correction (95% CrI: `r cri_pct(c2_tc$pct_lo, c2_tc$pct_hi)`). Vocal and postural effort followed the same pattern at roughly half the magnitude. Cumulative vocal effort increased by `r pct(c1_ec$pct)` at the first correction (95% CrI: `r cri_pct(c1_ec$pct_lo, c1_ec$pct_hi)`) and by `r pct(c2_ec$pct)` at the second (95% CrI: `r cri_pct(c2_ec$pct_lo, c2_ec$pct_hi)`), and cumulative postural effort increased by `r pct(c1_cc$pct)` at the first correction (95% CrI: `r cri_pct(c1_cc$pct_lo, c1_cc$pct_hi)`) and by `r pct(c2_cc$pct)` at the second (95% CrI: `r cri_pct(c2_cc$pct_lo, c2_cc$pct_hi)`). 

To test whether these cumulative increases reflected more movement or larger movements, we refitted the models with trial duration as a covariate. Adding duration substantially attenuated the correction effect (cf. Supplementary Material section D). The first-correction effect on the upper-limb effort remained credibly positive, but decreased from `r pct(c1_tc$pct)` to `r pct(c1_tc_wt$pct)` (95% CrI: `r cri_pct(c1_tc_wt$pct_lo, c1_tc_wt$pct_hi)`), suggesting that almost two-thirds of the increase in upper-limb effort can be attributed to accumulation of more, rather than larger, arm movement. The second-correction effect  on upper-limb effort and both corrections effects for vocal and postural effort, were fully accounted for by duration: with duration in the model, none remained credibly different from zero. 

None of the instantaneous measures showed a similar credible pattern: at the populational level, the instantaneous vocal, upper-limb, and postural effort remained unchanged following a misunderstanding. In an exploratory analysis, however, a model including the interaction between correction and each performer’s average baseline (i.e., first performance) effort showed that the first-correction effect is credibly influenced by the amount of effort in the baseline performance, and can even lead in the opposite direction (see Supplementary Material section F, Fig X). Specifically, for instantaneous arm torque change, the interaction with baseline effort was strongly negative: performers with low average baseline effort (−2 SD from the mean) increased their first correction by `r pct(int_in_torq_lo$pct)` (95% CrI: `r cri_pct(int_in_torq_lo$pct_lo, int_in_torq_lo$pct_hi)`), while those with high baseline effort (+2 SD) decreased it by `r pct(int_in_torq_hi$pct)` (95% CrI: `r cri_pct(int_in_torq_hi$pct_lo, int_in_torq_hi$pct_hi)`). Instantaneous vocal effort showed the same dependance: low-baseline performers (-2 SD) increased their first correction by `r pct(int_in_env_lo$pct)` (95% CrI: `r cri_pct(int_in_env_lo$pct_lo, int_in_env_lo$pct_hi)`), while high-baseline performers (+2 SD) decreased it by `r pct(int_in_env_hi$pct)` (95% CrI: `r cri_pct(int_in_env_hi$pct_lo, int_in_env_hi$pct_hi)`). For postural effort, the pattern was weaker: high-baseline performers decreased their first correction by `r pct(int_in_cop_hi$pct)` (95% CrI: `r cri_pct(int_in_cop_hi$pct_lo, int_in_cop_hi$pct_hi)`), while low-baseline performers showed no credible change `r fmt_fe_pct(int_in_cop_lo)`. Instantaneous effort at correction was therefore not uniformly recruited; whether performers scaled up the instantaneous effort depended on how effortful their baseline performance already was. This pattern warrants caution for two reasons. First, the baseline-by-correction interaction is susceptible to regression to the mean: participants with ‘extreme’ behavior will tend to regress toward the group mean at subsequent timepoints, which could artifactually produce the observed sign reversal. Second, high-baseline performers may face ceiling constraints on instantaneous effort, making decrease more likely by default. Because the present analyses cannot distinguish between statistical artifacts and behavioral patterns, we do not interpret it substantively. 

We also asked how baseline effort levels differed across modality conditions, since the available articulators differ between them. The modality shaped effort, but in measure-specific ways. Vocal effort was substantially higher in the vocal-only than in the multimodal condition, both cumulatively (`r pct(mod_ec$pct)` higher, 95% CrI `r cri_pct(mod_ec$pct_lo, mod_ec$pct_hi)`) and instantaneously (`r pct(mod_ei$pct)` higher, 95% CrI `r cri_pct(mod_ei$pct_lo, mod_ei$pct_hi)`), implying that vocal effort takes on more of the communicative burden when the hands are unavailable. The effect of modality on postural effort is reversed in direction: conditions involving bodily movement elicited far greater COPc than the vocal-only condition, both cumulatively (gesture-only: `r pct(go_vs_vo_cc$pct)` higher, 95% CrI: `r cri_pct(go_vs_vo_cc$pct_lo, go_vs_vo_cc$pct_hi)`; multimodal: `r pct(mm_vs_vo_cc$pct)` higher, 95% CrI: `r cri_pct(mm_vs_vo_cc$pct_lo, mm_vs_vo_cc$pct_hi)`) and instantaneously (gesture-only: `r pct(go_vs_vo_ci$pct)` higher, 95% CrI: `r cri_pct(go_vs_vo_ci$pct_lo, go_vs_vo_ci$pct_hi)`, multimodal: `r pct(mm_vs_vo_ci$pct)` higher,
95% CrI: `r cri_pct(mm_vs_vo_ci$pct_lo, mm_vs_vo_ci$pct_hi)`). Gesture-only and multimodal conditions did not differ in cumulative postural effort, although multimodal exceeded gesture-only for instantaneous postural effort, suggesting that coordinating two modalities involves bursts of postural activity without increasing total postural load. For upper-limb effort, we found no reliable average difference between conditions once between-participant and between-concept variability was modelled. The gesture-only vs. multimodal contrast was positive for cumulative effort but its interval included zero (`r pct(mod_tc$pct)`, 95% CrI `r cri_pct(mod_tc$pct_lo, mod_tc$pct_hi)`). Substantial random slope variance indicated that who is communicating and which concept is conveyed matter more for upper-limb effort than the modality condition itself.

To summarize, while cumulative effort increased reliably across corrections for all three measures, instantaneous effort did not. Correction therefore imposes a clear cumulative cost on performers without recruiting greater peak intensity per movement or per vocalization. This pattern is consistent with a redistribution of effort across longer time intervals and more movement units rather than a concentrated scaling-up of force or amplitude, an interpretation we develop in the Discussion. The consistency of these increases across three independent measures of bodily effort supports that repair imposes a substantive physical cost on the performer, supporting H1 for cumulative, though not instantaneous, effort.

![**Fig. 1: Task design and empirical results.** (A) Experimental design and measurements. A performer expresses a target concept to a guesser under one of three modality conditions (gesture-only, vocal-only, or multimodal); after an incorrect response, the performer produces a correction, up to two corrections per concept (baseline guess ‘journey’, incorrect; correction guess ‘to walk’, correct). We extract three effort signals: torque change of the arms (upper-limb effort), acoustic amplitude envelope (vocal effort), and change in center of pressure (postural effort), each summarized as cumulative effort (area under the curve) and instantaneous effort (mean of detected peaks). These give the six effort variables used throughout. (B) Posterior distribution of effort increase across corrections, on the log scale: first correction (left panel) is relative to baseline, and second correction (right panel) is relative to first correction. Dashed line at 0: no change; point: posterior median; thick and thin lines: 89% and 95% highest density intervals (HDI). (C) Posterior distribution of effort increase per 1 SD of cosine similarity between the answer and the target meaning (%, see Methods); higher similarity reflects less misunderstanding. Conventions as in B. (D) Posterior slopes relating the six effort variables to four measures of communicative success, each a separate model (see Methods): probability of understanding at the first-attempt (log-odds); answer similarity to target (Beta scale); understanding across attempts (ordinal: first attempt < first correction < second correction < unresolved; log-odds); and whether ever resolved (log-odds). Conventions as in B.](/../assets/Effort-multiplot.png)

## Upper-limb effort scales with the degree of misunderstanding

```{r}
#| label: setup-prev-answer
#| include: false


h2m_ct <- readRDS(here::here("06_Analysis_Modeling/models/h2.m1b_torq_cum.rds"))
h2m_it <- readRDS(here::here("06_Analysis_Modeling/models/h2.m1b_torq_in.rds"))

prev_an_ct_ci <- fe(h2m_ct, "answer_prev_dist_z")
prev_an_it_ci <- fe(h2m_it, "answer_prev_dist_z")

```

We found partial support for H2 (Fig. 2c). Cumulative upper-limb effort scaled credibly with the degree of misunderstanding, operationalized as the cosine similarity between the guesser’s previous answer and the target meaning (lower similarity indicates greater misunderstanding). Each 1-SD increase in cosine similarity was associated with `r pct(prev_an_ct_ci$pct)` reduction in cumulative arm torque change (95% CrI: `r cri_pct(prev_an_ct_ci$pct_lo, prev_an_ct_ci$pct_hi)`); equivalently, performers invested more cumulative upper-limb effort when the previous answer was further from the target. Instantaneous arm torque change ran in the same direction, but its estimated interval included zero `r fmt_fe_pct(prev_an_it_ci)`. In contrast, neither cumulative nor instantaneous vocal effort showed a credible relationship with cosine similarity, and the same was true for both postural measures.

Cumulative upper-limb effort therefore adapts reliably to the amount of repair work the previous answer requires, while vocal and postural effort appear insensitive to the degree of misunderstanding. The asymmetry between cumulative and instantaneous upper-limb effort mirrors the pattern reported in the previous subsection: misunderstanding scales cumulative effort more reliably than instantaneous effort.

Importantly, the effect of semantic distance on cumulative upper-limb effort was uniformly distributed across the sample: the random-slope variance for cosine similarity was negligible across all three grouping levels (concepts, participants, and dyads), indicating that the effect did not vary systematically by performer, dyad, or concept. This consistency suggests that arm movement responds to common ground in a relatively automatic way, rather than reflecting strategic adaptation by particular individuals or for particular concepts.

## Effort does not predict communicative success


```{r extract-all, include=FALSE}

# ── Models ────────────────────────────────────────────────────────────────────
e3_m1        <- readRDS(here::here("06_Analysis_Modeling/models/e3.m1.rds"))
e3_m1b       <- readRDS(here::here("06_Analysis_Modeling/models/e3.m1b.rds"))
fit_sim      <- readRDS(here::here("06_Analysis_Modeling/models/e3.m3_att.rds"))
fit_sim_inst <- readRDS(here::here("06_Analysis_Modeling/models/e3.m3b_att.rds"))
fit_ord      <- readRDS(here::here("06_Analysis_Modeling/models/e3.m5_att.rds"))
fit_ord_inst <- readRDS(here::here("06_Analysis_Modeling/models/e3.m5b_att.rds"))
fit_ever     <- readRDS(here::here("06_Analysis_Modeling/models/e3.m6_att.rds"))
fit_ever_inst <- readRDS(here::here("06_Analysis_Modeling/models/e3.m6b_att.rds"))

CAT_C0    <- 1
CAT_NEVER <- 4

# ── First-attempt resolution (cumulative) ─────────────────────────────────────
nd_base      <- make_nd_base(e3_m1)
epred_base   <- posterior_epred(e3_m1, newdata = nd_base, re_formula = NA)
idx          <- setNames(seq_along(nd_base$modality), nd_base$modality)

p_gest_base  <- ci(epred_base[, idx["gesture"]])
p_mm_base    <- ci(epred_base[, idx["multimodal"]])
p_voc_base   <- ci(epred_base[, idx["vocal"]])

dp_gest_vs_mm <- ci(epred_base[, idx["gesture"]]    - epred_base[, idx["multimodal"]])
dp_voc_vs_mm  <- ci(epred_base[, idx["vocal"]]      - epred_base[, idx["multimodal"]])

# ── Answer similarity (cumulative) ────────────────────────────────────────────
nd_base_sim <- make_nd_base(
  fit_sim,
  extra_zero_vars   = "answer_prev_dist_z",
  extra_factor_refs = list(correction = levels(fit_sim$data$correction)[1])
)
epred_base_sim <- posterior_epred(fit_sim, newdata = nd_base_sim, re_formula = NA)
idx_sim        <- setNames(seq_along(nd_base_sim$modality), nd_base_sim$modality)

p_gest_sim_base <- ci(epred_base_sim[, idx_sim["gesture"]])
p_mm_sim_base   <- ci(epred_base_sim[, idx_sim["multimodal"]])
p_voc_sim_base  <- ci(epred_base_sim[, idx_sim["vocal"]])

dp_gest_vs_mm_sim <- ci(epred_base_sim[, idx_sim["gesture"]]    - epred_base_sim[, idx_sim["multimodal"]])
dp_voc_vs_mm_sim  <- ci(epred_base_sim[, idx_sim["vocal"]]      - epred_base_sim[, idx_sim["multimodal"]])

# ── Resolution speed / ordinal (cumulative) ───────────────────────────────────
nd_base_ord <- make_nd_base(fit_ord)
epred_ord   <- posterior_epred(fit_ord, newdata = nd_base_ord, re_formula = NA)
idx_ord     <- setNames(seq_along(nd_base_ord$modality), nd_base_ord$modality)

pc0_gest_ord  <- ci(ord_cat(epred_ord, idx_ord["gesture"],    CAT_C0))
pc0_mm_ord    <- ci(ord_cat(epred_ord, idx_ord["multimodal"], CAT_C0))
pc0_voc_ord   <- ci(ord_cat(epred_ord, idx_ord["vocal"],      CAT_C0))
pnev_mm_ord   <- ci(ord_cat(epred_ord, idx_ord["multimodal"], CAT_NEVER))
pnev_voc_ord  <- ci(ord_cat(epred_ord, idx_ord["vocal"],      CAT_NEVER))

dp_c0_gest_vs_mm_ord <- ci(ord_cat(epred_ord, idx_ord["gesture"],    CAT_C0) -
                            ord_cat(epred_ord, idx_ord["multimodal"], CAT_C0))
dp_c0_voc_vs_mm_ord  <- ci(ord_cat(epred_ord, idx_ord["vocal"],      CAT_C0) -
                            ord_cat(epred_ord, idx_ord["multimodal"], CAT_C0))

# Credible interaction: envelope × vocal (instantaneous ordinal)
fe_ord_inst          <- fixef(fit_ord_inst, summary = FALSE)
nd_base_ord_inst     <- make_nd_base_in(fit_ord_inst)
epred_ord_inst       <- posterior_epred(fit_ord_inst, newdata = nd_base_ord_inst, re_formula = NA)

b_env_x_voc_ord_inst      <- ci(fe_ord_inst[, "modality3:envelope_in_log_c"])
prod_c0_env_voc_ord_inst  <- productivity_ord(fit_ord_inst, nd_base_ord_inst, "vocal", "envelope_in_log_c", CAT_C0)
prod_nev_env_voc_ord_inst <- productivity_ord(fit_ord_inst, nd_base_ord_inst, "vocal", "envelope_in_log_c", CAT_NEVER)

# ── Ever resolved ─────────────────────────────────────────────────────────────
nd_base_ever <- make_nd_base(fit_ever)
epred_ever   <- posterior_epred(fit_ever, newdata = nd_base_ever, re_formula = NA)
idx_ever     <- setNames(seq_along(nd_base_ever$modality), nd_base_ever$modality)

p_gest_ever <- ci(epred_ever[, idx_ever["gesture"]])
p_mm_ever   <- ci(epred_ever[, idx_ever["multimodal"]])
p_voc_ever  <- ci(epred_ever[, idx_ever["vocal"]])

dp_gest_vs_mm_ever <- ci(epred_ever[, idx_ever["gesture"]]    - epred_ever[, idx_ever["multimodal"]])
dp_voc_vs_mm_ever  <- ci(epred_ever[, idx_ever["vocal"]]      - epred_ever[, idx_ever["multimodal"]])

# Credible interactions (instantaneous ever-resolved)
nd_base_ever_inst  <- make_nd_base_in(fit_ever_inst)
epred_ever_inst    <- posterior_epred(fit_ever_inst, newdata = nd_base_ever_inst, re_formula = NA)
idx_ever_inst      <- setNames(seq_along(nd_base_ever_inst$modality), nd_base_ever_inst$modality)

fe_ever_inst           <- fixef(fit_ever_inst, summary = FALSE)
b_env_x_mm_ever_inst   <- ci(fe_ever_inst[, "modality1:envelope_in_log_c"])
b_copc_x_voc_ever_inst <- ci(fe_ever_inst[, "modality3:copc_in_log_c"])

prod_ever_inst_env_mm   <- productivity(fit_ever_inst, nd_base_ever_inst, "multimodal", "envelope_in_log_c")
prod_ever_inst_copc_voc <- productivity(fit_ever_inst, nd_base_ever_inst, "vocal",      "copc_in_log_c")

p_voc_ever_inst_copc_lo <- ci(pred_at(fit_ever_inst, nd_base_ever_inst, "vocal", "copc_in_log_c", -1))
p_voc_ever_inst_copc_hi <- ci(pred_at(fit_ever_inst, nd_base_ever_inst, "vocal", "copc_in_log_c",  1))
```

In an exploratory analysis, we asked whether the effort performers invested in repair translated into improved communicative outcomes. Across four complementary outcomes (1. whether the misunderstanding was resolved on the first attempt, 2. how semantically similar the guesser’s answer was to the target concept, 3. how quickly understanding was reached across correction attempts, and 4. whether understanding was ever achieved across the full sequence), modality and expressibility were the dominant predictors and effort was not (Fig. 1d). 

Multimodal trials outperformed vocal-only trials across all outcomes. The predicted probability of reaching understanding at first-attempt was `r fmt_p(p_mm_base["med"], p_mm_base["lo"], p_mm_base["hi"])` for multimodal
and `r fmt_p(p_voc_base["med"], p_voc_base["lo"], p_voc_base["hi"])` for vocal-only (contrast = `r fmt_dp(dp_voc_vs_mm["med"])`, 95% CrI [`r fmt_dp(dp_voc_vs_mm["lo"])`, `r fmt_dp(dp_voc_vs_mm["hi"])`]), and the probability of never reaching understanding was (P(never) = `r fmt_p(pnev_mm_ord["med"], pnev_mm_ord["lo"], pnev_mm_ord["hi"])`
for multimodal and `r fmt_p(pnev_voc_ord["med"], pnev_voc_ord["lo"], pnev_voc_ord["hi"])` for vocal-only). Gesture-only performances were largely indistinguishable in terms of success measures from multimodal although it tended in the same direction as vocal-only trials. Similarly, gesture-only performances did not reliably differ from vocal-only performances, although the distribution tended towards the gesture-only condition yielding better success.

Despite the clear effort increase during repair, this investment does not translate into communicative gain. Effort slopes were small, inconsistent in direction, and non-credible across the large majority of effort × modality combinations. A small number of exceptions ran in opposing directions and warrant cautious interpretation. Instantaneous vocal amplitude predicted faster understanding in vocal-only trials, and greater postural adjustments at first attempt predicted higher probability of eventually reaching understanding in vocal-only trials, possibly indexing speaker commitment when gesture is unavailable. In the opposite direction, higher instantaneous vocal effort in multimodal trials was associated with lower probability of eventual understanding, consistent with louder vocalization marking difficulty rather than driving success.

# Discussion
When people fail to communicate a concept through non-linguistic means, they do not simply reorganize their behavior – they invest more physical effort. And they do so in a specific way: not by doing the same thing harder, but by doing more over time. This distinction between intensification and elaboration may reflect a general principle of how the body responds to misunderstanding: effort expands across a longer temporal window rather than concentrating locally. Our data suggest this response is also flexible – sensitive to how large the misunderstanding is and shaped by the articulatory resources available. We thereby show that misunderstanding literally recruits investment of physical (metabolic) cost, thereby providing a fundamental grounding of “cost” that has so far remained implicit in accounts of effort and efficiency in linguistics.

These findings extend existing accounts of how environmental constraints reshape communicative behavior. Lindblom’s H\&H Theory [@lindblomExplainingPhoneticVariation1990] predicts that a system under pressure produces a clearer, more precise signal. Rather than contradicting this prediction, our results reveal that misunderstanding imposes a categorically different kind of demand than, say, a noisy room. Situations with background noise may require a local enhancement (e.g., louder speech) relative to the environment. Epistemic misalignment, on the other hand, necessitates enhancement that is more global: adding more information and changing the structure of the message. This is in line with existing work on communicative repair [@rasenbergMultimodalNatureCommunicative2022; @hilliardBridgingGapsCommon2016] but our operationalization allows us to claim, for the first time, that performers invest more physical effort rather than simply reorganizing the content of the message when their addressee signals poor understanding.

Crucially, elaboration is not reflected in the same manner across dimensions. Upper-limb movement shows the greatest capacity for cumulative increase: gesture offers multiple routes through which more structure can be added such as pointing, tracing, representing, and the increase may also reflect cognitive engagement with the target, given gesture's link to conceptual and lexical processing [@hostetterSeeItMy2007; @kraussRoleSpeechrelatedArm1999]. VVocal effort showed a more modest increase, possibly reflecting the limited degrees of freedom available for non-linguistic vocalization: once referential breakdown has occurred, the semantic space is harder to navigate without the semiotic flexibility that gesture affords. Postural effort also increased cumulatively across conditions; its smaller magnitude is consistent with a stability-oriented function [@massionMovementPostureEquilibrium1992] and its presence even in the vocal-only condition suggests that misunderstanding imposes non-representational demands on the whole body, possibly reflecting cognitive load or self-regulatory mechanisms [@yardleyEffectArticulatoryMental1999; @linStandTallThink2026].

The diverging articulatory and meaning-making affordances in gestural and vocal modality hint that the two modalities complement each other, though our data show this complementarity is asymmetric. Vocal effort was substantially lower in the multimodal than in the vocal-only condition, indicating that the voice carries less of the communicative load when the upper limbs are available. The complementary comparison – upper-limb effort in multimodal versus gesture-only condition – pointed in the same direction (smaller upper-limb effort when voice is present) but its interval included zero. With respect to communicative success, gesture and vocalization together outperformed vocalization alone, but not gesture alone across all four outcomes. Together, this implies that adding gesture to voice helps more than adding voice to gesture, further corroborating the semiotic flexibility that gesture affords. Once expressibility is accounted for, we found no credible difference in communicative success between gesture-only and vocal-only trials, despite the gestural advantage  reported in the literature [@fayCreatingCommunicationSystem2014; @fayGesturePrimaryModality2022]. Unlike prior work, our model includes expressibility, which was by far the strongest driver of success. Because expressibility captures how well a concept’s affordances match a given modality, it likely absorbs the variance that previously appeared as a gesture-vocal difference – extending with our earlier work on guessability in this dataset (@cwiekSelfreportedExpressibilityPredictsaccepted). The modality contrast, in other words, may reflect concept-modality fit rather than an intrinsic advantage of gesture over vocalization. 

The disconnect between effort and communicative success is clearest here: although performers recruited more effort under misunderstanding, that effort did not lead to better outcomes. Such decoupling is not uncommon [@wolffWhenEffortFails2025a] and the available evidence on this question is mixed. While some work hints at greater effort bringing greater success [@drijversVisualContextEnhanced2017], many studies leave the relationship unresolved [@rasenbergMultimodalNatureCommunicative2022; @trujilloSpeakersExhibitMultimodal2021; @hollerExperimentalInvestigationHow2011] or fail to find it altogether [@bogelsTherePowerLaw2024]. Our data indicate that repair may be a context where effort serves a different function altogether. In a noisy room, a strong, clear signal is a precondition for any interaction to take place at all. Misunderstanding may call for something else. Instead of more effort, the interlocutor could benefit from change of tempo, greater informativity of individual units, or – as our data show – the number of communicative channels available. The persistent effect of expressibility on both effort and success in our data points in the same direction: a specific meaning shapes how effortful it is to communicate and how readily it is understood, somewhat independent of how hard the performer works. What does seem clear is that the communicative value of multimodality lies not in effort efficiency but in the richer articulatory palette it affords. 

Yet effort is not decoupled from the communicative context altogether. Our second pre-registered analysis shows that the relationship runs in the other direction: rather than driving the communicative outcome, effort is a response to it. Cumulative upper-limb effort scaled with the degree of misunderstanding, decreasing as the guesser’s previous answer approached the target meaning. This calibration was specific to the arms; neither vocal nor postural effort tracked the degree of misunderstanding. It was also uniform across the sample, varying little across performers, dyads, or concepts, which suggests that the adjustment of arm effort to common ground operates similarly across individuals rather than reflecting idiosyncratic strategy. Similarly to the correction effect in H1, instantaneous measures were largely unmoved by misunderstanding. The consistent picture is that misunderstanding reshapes how long and how much performers communicate, not how forcefully. This is in support of audience design accounts [@CLARK1982287; @hortonWhenSpeakersTake1996] proposing that people monitor the current state of shared understanding and tune their behavior to their partners when necessary. Future work shall determine the qualitative nature of this elaboration, exploring the relationship between different types of misunderstanding and the attunements they trigger.

Our findings must be interpreted in light of several limitations. The controlled design of our paradigm, while necessary for isolating effort modulation, comes with trade-offs in ecological validity. Whole-body movement without speech amplifies physical demands beyond what typical face-to-face interaction requires, and in naturalistic settings linguistic structure carries much of the repair work. We view this as a necessary first step: charade-like referential tasks share fundamental properties with linguistic interaction: they are inherently communicative, meaningful, and intersubjective [@christiansenLanguageGameHow2022]. Identifying effort modulation in this simplified context establishes a foundation for investigating these phenomena in more naturalistic settings.

A related limitation concerns what counts as communicative success. Our design treats understanding as an exact match between the guesser's answer and the target concept, whereas real interaction typically requires only sufficient mutual understanding to support the next action. More consequentially, our paradigm places most of the repair work on the performer: the guesser contributes only a single guess per performance, leaving little room to act on the performer's increased effort. Research on referential tasks with freer interaction suggests that more distributed repair (where partners jointly signal and respond to misunderstanding) may be what drives communicative success [@rasenbergMultimodalNatureCommunicative2022]. It is therefore possible that effort does not help in our paradigm precisely because the partner who could benefit from it has limited room to act on it. 

A second set of limitations concerns the measurement of effort itself. The three features we selected – arm torque, acoustic amplitude envelope, and centre of pressure displacement – capture general biomechanical demands on the body, but effort is content-dependent: concepts vary substantially in how much and what kind of effort their communication requires, as reflected in the large concept-level variance in our models. While our population-level effects suggest that the selected features capture a general mechanism, future work should examine whether other dimensions of movement such as speed, spatial precision, or inter-articulator coordination carry additional information that our measures do not. In particular, the coordinative cost of orchestrating multiple articulators simultaneously remains unaddressed here and may represent a distinct dimension of multimodal effort with its own scaling properties. Our preregistration [@kadavaPuttingEffortModulation2025] includes an example of such exploratory analysis, which we will address in an upcoming work. Taken together, these limitations point toward productive directions for future research without undermining the core finding: that the body responds to misunderstanding with cumulative, distributed effort reliably, across modalities, and, for the arms, in proportion to the degree of communicative failure. 

Together, the absence of an effort-success link and the responsiveness of effort to feedback point toward a reframing of what communicative effort represents. Rather than a resource allocated toward success, effort appears to function as a continuous index of the interaction's epistemic state: signaling bodily mobilization when understanding breaks down, scaling with the distance from common ground, and recalibrating when that distance closes. These findings  sit in tension with accounts of principle of least effort and effort aversion [@andrePeopleReallyAvoid2026; @jobChallengingLawLeast2024]. If people generally avoid effortful behavior [@davidUnpleasantnessThinkingMetaanalytic2024], why would they increase physical investment in response to misunderstanding without clear returns in success? The standard evolutionary view holds that efficient signals convey as much as possible at minimal cost but our findings indicate a complementary pressure: effort may be organized not around maximizing informational yield but around sustaining the communicative cycle itself. What costs something is not saying more but staying in the exchange long enough to be understood. Research on challenging tasks suggests that this kind of investment is not irrational: effort carries intrinsic social value, enhancing the perceived worth of its product including fostering social belonging [@inzlichtEffortParadoxEffort2018]. On this view, the effort enhancement we observe may be less about solving the problem and more about affirming commitment to the partner at the moment the interaction is most at risk of failing.

The relational function of effort may extend beyond humans. When animals repeat and vary their signals in the face of failure, this is taken as one of the objective criteria of intentional communication [@townsendExorcisingGricesGhost2017]. Our findings suggest that intentionality may leave a measurable physical trace, as sustained effort as operationalized here must carry a metabolic cost. What happens to that trace when breakdown escalates into conflict, when the interaction is particularly vulnerable, or when competing demands leave fewer bodily resources available? These are the conditions where the effort to stay in it together – for humans and animals alike – matters most. 
