---
title: "Untitled"
format:
  html:
    code-overflow: wrap
    code-width: 1200  # Adjust the width in pixels
execute:
   warning: false 
toc: true
---

In this script, we are going to follow the pre-registered pipeline for modelling the relationship of effort and corrections. to confirm/reject our hypothesis.

These are:

H1: In corrections, people enhance some effort-related kinematic and/or acoustic features of their behaviour relative to the baseline.

H2: The enhancement depends on similarity of the guesser's answer and the original meaning. More similar answer will require/result in smaller enhancement (but still enhancement) than less similar answer.

The confirmatory analysis concerns six outcome variables, namely integral of torque change, integral of amplitude envelope, integral of change in center of pressure, and mean peak value of torque change, mean peak value of amplitude envelope, and mean peak value of change in center of pressure.

```{r setup environment, echo=TRUE, message=FALSE, warning=FALSE}
#| code-fold: true
#| code-summary: Code to prepare the environment

library(here)
library(dplyr) # for data-wrangling
library(tidyr)  # for reshaping data (if needed)
library(ggplot2)
library(tibble)
library(rcartocolor)
library(patchwork)
library(rstan)
library(ggdag) # for dag
library(dagitty)

library(bayesplot) # bayesian stuff
library(brms)
options(mc.cores = 4, brms.backend = 'cmdstanr', brms.file_refit = 'on_change')

library(beepr)
library(bayestestR)
library(tidyverse)

library(splines) # for bsplines
library(emmeans) # for lognormal transformations


# current folder (first go to session -> set working directory -> to source file location)
curfolder     <- getwd()
parentfolder  <- dirname(curfolder)
featurefolder <- paste0(parentfolder, '/07_TS_featureExtraction/Datasets/')
datasets      <- paste0(curfolder, '/datasets/')
models        <- paste0(curfolder, '/models/')
plots         <- paste0(curfolder, '/plots/')


```


# Loading in our data

This is the feature dataframe we will be using. It contains all features colected in XX, together with demographic information about the participants.

```{r loading data, echo=TRUE, message=FALSE, warning=FALSE}

data_feat <- read_csv(paste0(featurefolder, "features_df_nonzscored_final_with_demo.csv"))
head(data_feat)

```

For confirmatory analysis, we will need only 6 features - arm torque integral, arm torque peak mean, COPc integral, COPc peak mean, envelope integral and envelope peak mean from the data_feat. From data_demo, we will need only BFI and Familiarity

```{r picking columns, echo=TRUE, message=FALSE, warning=FALSE}

featstokeep <- c('COPc_peak_mean', 'envelope_norm_peak_mean', 'arm_moment_sum_change_peak_mean', 'COPc_integral', 'envelope_norm_integral', 'arm_moment_sum_change_integral', 'correction_info', 'concept', 'TrialID', 'modality', 'expressibility', 'answer_prev', 'answer_prev_dist', 'pcn_ID', 'BFI_extra', 'BFI_agree', 'BFI_consc', 'BFI_negemo', 'BFI_open', 'Familiarity')

# Subset the dataframe
data_feat <- data_feat %>% select(all_of(featstokeep))

```

# H1: Stating causal model

Before simulating synthetic data, we will first formulate our causal model to explicitly identify common causes, and hence prevent (or rather minimize) confounding in our statistical model. Note that while we are interested in causal relationship between effort (outcome) and communicative attempt (predictor for H1) and answer similarity (predictor for H2), we need to acknowledge that there are other factors possibly contributing to the outcome variable, despite the experiment beling relatively controlled. These relate, for instance, to the participants (e.g., personality traits) or stimuli (i.e., concepts). Our causal model therefore includes more variables that are identified as potentially sharing causal path with our primary predictor - communicative attempt. They are not of primary interest, they ensure we can isolate the causal effect of the main predictor on the outcome variable - effort.

To explicitly formulate the causal paths, we will start our analysis visualizing our causal model using directed acyclic graph (DAG). We will use packages `dagitty` and `ggdag` to create these DAGs and to extract implied independencies and adjustment set. DAG will help us clarify the causal model and introduce the predictors we aim to model [@mcelreath18]. This model goes hand in hand with the logic underlying the synthetic data we create below.

Our first hypothesis goes as follows:

*H1: Correction recruits more physical effort than the baseline performance.*

The relationship of interest is the causal effect of *communicative attempt* on *effort*. Further, our causal model includes following assumptions:

-   Personality traits (measured with BFI) will influence effort (e.g., people are more willing to put more effort if they are open-minded) and also communicative attempt (e.g., more extroverted people are better in this game, therefore they need less attempts)

-   Familiarity with guessing partner influences effort (ditto) as well as communicative attempt (e.g., friends are better in this game than strangers)

-   Similarly, participant will also directly influence effort and commAtt, because personality traits might not be capturing all variation

-   Expressibility of concepts is going to influence effort (e.g., more expressible concepts allow more enhancement - but could be also other direction) as well as CommAtt (e.g., more expressible concepts are more readily guessed and don't need more attempts)

-   Similarly, concept will also directly influence effort and commAtt, because expressibility might not be capturing all variation

-   Modality (uni vs. multi) will influence the value of effort. We assume that in unimodal condition, the feature does not need to account for synergic relations with the other modality, and may carry the whole signal. In multimodal condition, however, there may be synergic relations that moderate the full expressiveness of this feature

-   Trial number (characterising how far one is in the experiment) could be hinting on learning processes through out the experiment, or - the other direction - on increasing fatigue

Note that we are aware that some of the parameters might be correlated and we might run into issues with collinearity. While DAG does not help to solve it, it does help in detecting whether the two potentially highly correlated variables have separate causal effect or share a common cause. If they share common cause, there might be a problem in distinguishing their effects from each other. We will report inter-variable correlations when analyzing the actual data.

```{r dag h1, echo=FALSE, message=FALSE, warning=FALSE}

daggy_h1 <- dagitty('dag {
Big5 [adjusted,pos="-1.5,1"]
CommAtt [exposure,pos="-2,0"]
Conc [adjusted,pos="-2,-1.5"]
Eff [outcome,pos="0,0"]
Expr [adjusted,pos="-1.5,-1.5"]
Fam [adjusted,pos="-0.5,1"]
Pcn [adjusted,pos="-1,2"]
TrNum [adjusted,pos="-2.5,-1.5"]
Mod [adjusted,pos="0.5,0.5"]

Big5 -> CommAtt     
Big5 -> Eff
CommAtt -> Eff
Conc -> Expr
Expr -> CommAtt
Expr -> Eff
Fam -> CommAtt
Fam -> Eff
Mod -> Eff
Mod -> CommAtt
Pcn -> Big5
Pcn -> CommAtt 
Pcn -> Eff
Pcn -> Fam
TrNum -> CommAtt
TrNum -> Eff
Conc -> CommAtt
Conc -> Eff
}')

# Use ggdag for a cleaner visualization
d1 <- ggdag(daggy_h1, text = TRUE, use_labels = "name") +
  theme_dag() +
  ggtitle("Causal DAG for H1") +
  theme(plot.title = element_text(hjust = 0.5))

d1

```

These are the implied conditional independencies

```{r dag h1 independencies, echo=FALSE, message=FALSE, warning=FALSE}

impliedConditionalIndependencies(daggy_h1)

```

This is the adjustment set that needs to be included in the model to make sure we are not confounding

```{r dag h1 adjustset, echo=FALSE, message=FALSE, warning=FALSE}

dagitty::adjustmentSets(daggy_h1, exposure = "CommAtt", outcome = "Eff")

```

# H1: Modelling

In this part, we are going to test the following hypothesis:

*H1: Correction recruits more physical effort than the baseline performance.*

We will fit Bayesian mixed effect models using `brms` package [@burkner17].

## Data wrangling

To be able to interpret the data, we will first need to do some data wrangling

Convert columns to factors

```{r h1 factorizing, message=FALSE, warning=FALSE}

final_data <- data_feat

# Create column TrialNumber by extracting third element from TrialID
final_data$TrialNumber <- sapply(strsplit(final_data$TrialID, "_"), function(x) x[3])
final_data$TrialNumber <- as.numeric(final_data$TrialNumber)  # Ensure TrialNumber is numeric

# Create sessionID by taking first element
final_data$SessionID <- sapply(strsplit(final_data$TrialID, "_"), function(x) x[1])
final_data$SessionID <- as.factor(final_data$SessionID)

# The rest as factor
final_data$correction_info <- as.factor(final_data$correction_info)
final_data$modality <- as.factor(final_data$modality)
final_data$pcn_ID <- as.factor(final_data$pcn_ID)
final_data$concept <- as.factor(final_data$concept)

```

Contrast-coding of categorical variables

```{r h1 contrast coding, message=FALSE, warning=FALSE}

final_data$correction_info <- factor(final_data$correction_info, 
                                   levels = c('c0_only', 'c0', 'c1', 'c2'))

# FLAG: Still not sure whether c0_only should be included in analysis but am hesitant to get rid of almost half of the data?


contrasts(final_data$correction_info) <- MASS::contr.sdif(4) 
#contrasts(final_data$modality) <- contr.sum(3)/2 # we will CC modality for each model separately

```

::: callout-note
This is how CommAtt is contrast-coded now

2-1  3-2   4-3
c0_only -0.75 -0.5 -0.25
c0       0.25 -0.5 -0.25
c1       0.25  0.5 -0.25
c2       0.25  0.5  0.75

This should be enough given that we are interested in the incrementals.
:::

Centering continuous variables


```{r h1 centering, message=FALSE, warning=FALSE}

final_data$TrialNumber_c <- final_data$TrialNumber - median(range(final_data$TrialNumber))
#final_data$Familiarity <- final_data$Familiarity - median(range(final_data$Familiarity))

# Z-score familiarity to be compatible with BFI
final_data$Familiarity <- scale(final_data$Familiarity, center = TRUE, scale = TRUE) |> as.numeric()

# FLAG: Not sure here about Familiarity. First, do I even treat it correctly (ie it is just sum of two questions), second: should it then be centered or z-scored


```

Note: BFI has been z-scored when processing metadata as this is the common procedure for calculating each trait's score

Z-scoring expressibility within a modality

```{r h1 zscoring, message=FALSE, warning=FALSE}

final_data <-
  final_data |>
  group_by(modality) |>
  mutate(expressibility_z = (expressibility - mean(expressibility))/ sd(final_data$expressibility, na.rm = T)) |>
  ungroup()

```

## Var 1: Cumulative torque
### Model 0 - Intercept only

```{r h1m0 data wrangling, echo=TRUE, message=FALSE, warning=FALSE}

# Get rid of vocal modality
df_ges <- final_data |> filter(modality != 'vocal')

# Dummy code modality
df_ges$Modality_dummy <- ifelse(df_ges$modality == "multimodal", 1, 0)
df_ges$Modality_dummy <- as.factor(df_ges$Modality_dummy)

```


```{r h1m0 torq cum, echo=TRUE, message=FALSE, warning=FALSE}


if (file.exists(here(models, "h1.m0_torq_cum.rds"))){
  
  h1.m0_torq_cum <- readRDS(here(models, "h1.m0_torq_cum.rds"))
  h1.m0_torq_cum_R2 <- readRDS(here(models, "h1.m0_torq_cum_R2.rds"))
  
} else {
  h1.m0_torq_cum <- brm(arm_moment_sum_change_integral ~ 1 + correction_info, 
                  data = df_ges,
                  family = lognormal(),
                  iter = 2000,
                  cores = 4)
  
  # Add criterions for later diagnostics
  h1.m0_torq_cum <- add_criterion(h1.m0_torq_cum, criterion = c("loo", "waic"))
  
  # Calculate also variance explained (R^2)
  h1.m0_torq_cum_R2 <- bayes_R2(h1.m0_torq_cum)
  
  # Save both as objects
  saveRDS(h1.m0_torq_cum, here(models, "h1.m0_torq_cum.rds"))
  saveRDS(h1.m0_torq_cum_R2, here(models, "h1.m0_torq_cum_R2.rds"))

}

# Summary
summary(h1.m0_torq_cum)

```
Check the diagnostics

```{r h1m0 diagnostics, echo=TRUE, message=FALSE, warning=FALSE}

plot(h1.m0_torq_cum)
# all look ok

plot(conditional_effects(h1.m0_torq_cum), points = TRUE)


pp_check(h1.m0_torq_cum, type = "dens_overlay")
# Very good

pp_check(h1.m0_torq_cum, type = "error_scatter_avg")
# The bigger the value, the bigger the residual error


```


### Model 1 - Simple reproduction of DAG 

First we want to do a simple model that reproduce our DAG.

Our main predictor is communicative attempt (CommAtt). To account for confounders - variables affecting both the predictor and the dependent variable - we need to adjust for them in the model by including them as covariates to isolate the effect of the predictor. Based on our DAG, confounders include:

1.  familiarity
2.  big5
3.  expressibility
4.  trial number
5.  modality
6.  concept
7.  participant

We include 1.-5. as fixed factors. For 6.-7., we include varying intercepts as we expect that each participant and concept may have they own baseline level of effort and thus allow for individual variation. Partial pooling is also beneficial in that extreme values (or categories will fewer datapoints) will be shrunk toward the overal average.

We will now not include PrevAn (previous answer) variable because we will need to do some further data-wrangling when building model for H2. That is mainly because PrevAn has some NA values, concretely for first attempts. Models would automatically exclude NA data, and we would therefore loose all effort data for attempt 1. For H1, however, we want to keep it there.

This is our first model without setting any priors, leaving them to default values


```{r h1.m1 torque cum model, eval=TRUE, message=FALSE, warning=FALSE}

if (file.exists(here(models, "h1.m1_torq_cum.rds"))){
  
  h1.m1_torq_cum <- readRDS(here(models, "h1.m1_torq_cum.rds"))
  h1.m1_torq_cum_R2 <- readRDS(here(models, "h1.m1_torq_cum_R2.rds"))
  
} else {

  h1.m1_torq_cum <- brm(arm_moment_sum_change_integral ~ 1 + correction_info + Familiarity + BFI_extra + expressibility_z + TrialNumber_c + Modality_dummy + (1 | pcn_ID) + (1 | SessionID) + (1 | concept),
                  data = df_ges,
                  family = lognormal(),
                  iter = 2000,
                  cores = 4)
  
  # Add criterions for later diagnostics
  h1.m1_torq_cum <- add_criterion(h1.m1_torq_cum, criterion = c("loo", "waic"))
  
  # Calculate also variance explained (R^2)
  h1.m1_torq_cum_R2 <- bayes_R2(h1.m1_torq_cum)
  
  # Save both as objects
  saveRDS(h1.m1_torq_cum, here("models", "h1.m1_torq_cum.rds"))
  saveRDS(h1.m1_torq_cum_R2, here("models", "h1.m1_torq_cum_R2.rds"))

}

# Summary
summary(h1.m1_torq_cum)


```

::: callout-note
Note on modality: Modality1 represents the baseline difference between combined and the other modalities, and Modality2 represents the difference between gesture and vocal compared to combined.
:::

::: callout-note
Expressibility is z-scored so we will not get to the coefficient in the same way but we can check the conditinal effects for checking whether it looks good
:::

Let's also check the visuals

```{r h1.m1 torq cum check, echo=TRUE, message=FALSE, warning=FALSE}

plot(h1.m1_torq_cum)
# b_Intercept looks a bit weird, but otherwise nothing alarming

plot(conditional_effects(h1.m1_torq_cum), points = TRUE)


pp_check(h1.m1_torq_cum, type = "dens_overlay")
# Looks ok

pp_check(h1.m1_torq_cum, type = "error_scatter_avg")
# half blobby, half correlated, so still some room for improvement
# positive correlation means that errors increase with predicted values. So the model does perform well for some range, but becomes less reliable with increase in the predicted values
# also the blob is centered around 0 which is good
# it could be we are ignoring some interaction terms or non-linearity (which we know we kind of do). Transformation could also help (e.g., log). Of course, we are also still not specifying any priors so let's not yet make it a disaster

h1.m1_torq_cum_R2
# explained variance around 50%

```

Overall, the perfomance is going in good direction

Let's have another model for comparison.

We can assume that participants and concept have not only different baselines of effort (varying intercept). The effect of CommAtt on effort might vary across them too, hence we can try to add varying slopes for them and see whether the diagnostics improves. We will also add TrialNumber as a varying intercept, because we expect variation between earlier and later performances (because of learning, or opposite, fatigue).


### Model 2 - Varying slopes and intercepts

```{r h1.m2 run, eval=TRUE, message=FALSE, warning=FALSE}


if (file.exists(here(models, "h1.m2_torq_cum.rds"))){
  
  h1.m2_torq_cum <- readRDS(here(models, "h1.m2_torq_cum.rds"))
  h1.m2_torq_cum_R2 <- readRDS(here(models, "h1.m2_torq_cum_R2.rds"))
  
} else {
  
  
  h1.m2_torq_cum <- brm(arm_moment_sum_change_integral ~ 1 + correction_info + Modality_dummy + BFI_extra + Familiarity + expressibility_z +  (1 + correction_info | pcn_ID)  + (1 | SessionID) + (1 + correction_info | concept) + (1 | TrialNumber_c), 
                  data = df_ges,
                  family = lognormal(),
                  iter = 2000,
                  cores = 4)
  
  # Add criterions for later diagnostics
  h1.m2_torq_cum <- add_criterion(h1.m2_torq_cum, criterion = c("loo", "waic"))
  
  # Calculate also variance explained (R^2)
  h1.m2_torq_cum_R2 <- bayes_R2(h1.m2_torq_cum)
  
  # Save both as objects
  saveRDS(h1.m2_torq_cum, here("models", "h1.m2_torq_cum.rds"))
  saveRDS(h1.m2_torq_cum_R2, here("models", "h1.m2_torq_cum_R2.rds"))

}

# Summary
summary(h1.m2_torq_cum)
# Coefficients remain mostly unchanged 
# There are still some Rhats > 1



```


```{r h1.m2 torq cum check, message=FALSE, warning=FALSE}

plot(h1.m2_torq_cum)
# all look pretty fuzzy

plot(conditional_effects(h1.m2_torq_cum), points = TRUE)

pp_check(h1.m2_torq_cum, type = "dens_overlay")
# looks good

pp_check(h1.m2_torq_cum, type = "error_scatter_avg")
# Blobby, but still positively correlated

h1.m2_torq_cum_R2
# explained variance around 51%

```

One of the reasons why we might be getting the divergent transition is because of the correlation between slopes and intercepts in the previous model. Let's now get rid of it


### Model 3 - No correlation coefficient

```{r h1.m3 run, message=FALSE, warning=FALSE}


if (file.exists(here(models, "h1.m3_torq_cum.rds"))){
  
  h1.m3_torq_cum <- readRDS(here(models, "h1.m3_torq_cum.rds"))
  h1.m3_torq_cum_R2 <- readRDS(here(models, "h1.m3_torq_cum_R2.rds"))
  
} else {

  h1.m3_torq_cum <- brm(arm_moment_sum_change_integral ~ 1 + correction_info + Modality_dummy + BFI_extra + Familiarity + expressibility_z +  (1 + correction_info || pcn_ID) + (1 || SessionID) + (1 + correction_info || concept) + (1 || TrialNumber_c), 
                  data = df_ges,
                  family = lognormal(),
                  iter = 2000,
                  cores = 4)
                  
  # Add criterions for later diagnostics
  h1.m3_torq_cum <- add_criterion(h1.m3_torq_cum, criterion = c("loo", "waic"))
  
  # Calculate also variance explained (R^2)
  h1.m3_torq_cum_R2 <- bayes_R2(h1.m3_torq_cum)
  
  # Save both as objects
  saveRDS(h1.m3_torq_cum, here("models", "h1.m3_torq_cum.rds"))
  saveRDS(h1.m3_torq_cum_R2, here("models", "h1.m3_torq_cum_R2.rds"))

}

# Summary
summary(h1.m3_torq_cum)
# Coefficients remain mostly unchanged 


```


```{r h1.m3 torq cum check, message=FALSE, warning=FALSE}

plot(h1.m3_torq_cum)
# all good

plot(conditional_effects(h1.m3_torq_cum), points = TRUE)


pp_check(h1.m3_torq_cum, type = "dens_overlay")


pp_check(h1.m3_torq_cum, type = "error_scatter_avg")
# unchanged

h1.m3_torq_cum_R2
# explained variance remains around 51%

```

### Model 3.1 - Adding priors

Now we will add some minimal priors, giving the model some reasonably informative, yet weak information.

These are the priors that were used for the last model

```{r prior summary h1m3}

prior_summary(h1.m3_torq_cum)

```
We can see that for all betas, the priors are flat which might be also what is causing the Rhat>1. We will put weakly informative priors to limit extreme values. Given that we have no strong opinion about the sigmas, we will update only those that were defaulted to flat priors, unless later it seems implausible.
We also set sigma because student-t(3,0,2.5) has heavy tails

We set our priors in the following way:

- **Intercept**: The geometric mean of arm moment for the reference group is most plausibly around exp(7) ≈ 1,100 raw units, but we allow considerable uncertainty — 95% of prior mass spans exp(3) to exp(11), i.e. roughly 20 to 60,000 raw units.

- **correction_info**: We have no prior expectation of whether corrections increase or decrease effort. But we allow each correction step to plausibly up to nearly triple (×2.7) or halve (×0.37) arm moment. Effects larger than that are considered implausible.

- **expressibility**, **familiarity** and **BFI**: A 1 SD change in expressibility is unlikely to change arm moment by more than ×1.6 in either direction. We expect expressibility to have a modest effect if any

- **Modality**: witching modality could plausibly up to nearly triple or halve arm moment. We allow a larger range here than for continuous predictors, reflecting genuine uncertainty about modality effects.

- Global sd: A default fallback for any grouping factor not specified individually. Units within a group are expected to differ by at most ~35%, with typical differences closer to 0–16%.

- sigma: Within a given combination of participant, concept and correction phase, individual observations are unlikely to deviate by more than ×1.6 above or below their cell mean. Residual spread is expected to be modest

- sd of **pcn_ID**: Participants differ from the population mean by at most ~60% in either direction (×0.6 to ×1.6) in arm moment. Allows meaningful individual differences without permitting implausibly huge person-level variation.

- sd of **concept**: Concepts differ from the population mean by at most ~60% in either direction. Some concepts may elicit more arm effort than others, but not by an order of magnitude

- sd of **trial number**: Trial-to-trial drift in arm moment is expected to be very small — most likely within ±22% of the mean (exp(±0.2)), reflecting a minor practice or fatigue trend rather than a large systematic shift.


```{r priors for h1}

priors_h1m3p_torq_cum <- c(
  set_prior("normal(5, 2)", class = "Intercept"),
  set_prior("normal(0,0.25)", class = "b", coef = "BFI_extra"),
  set_prior("normal(0,0.5)", class = "b", coef = "correction_info2M1"),
  set_prior("normal(0,0.5)", class = "b", coef = "correction_info3M2"),
  set_prior("normal(0,0.5)", class = "b", coef = "correction_info4M3"),
  set_prior("normal(0,0.35)", class = "b", coef = "expressibility_z"),
  set_prior("normal(0,0.25)", class = "b", coef = "Familiarity"),
  set_prior("normal(0,0.5)", class = "b", coef = "Modality_dummy1"),
  set_prior("normal(0,0.15)", class = "sd"),
  set_prior("normal(0, 0.5)", class = "sigma"),
  # Random effect SDs: differentiated by grouping factor
  # Participants: largest expected source of variation
  set_prior("normal(0, 0.35)", class = "sd", group = "pcn_ID"),
  # Concepts: moderate
  set_prior("normal(0, 0.35)", class = "sd", group = "concept"),
  # Trial number: small drift
  set_prior("normal(0, 0.1)", class = "sd", group = "TrialNumber_c")
)
  

```

Now we do prior predictive check

```{r h1 prior check}

m_prior_check <- brm(
  arm_moment_sum_change_integral ~ 1 + correction_info + Modality_dummy + 
    BFI_extra + Familiarity + expressibility_z +  
    (1 + correction_info || pcn_ID) + (1 || SessionID) + 
    (1 + correction_info || concept) + (1 || TrialNumber_c),
  data = df_ges,
  family = lognormal(),
  prior = priors_h1m3p_torq_cum,
  sample_prior = "only",
  iter = 2000,
  cores = 4
)

pp_check(m_prior_check, ndraws = 1000)
```

There are some astronomical values that prevent us to actually see the data, does it help if we log the scale?

```{r h1 ppcheck}
pp_check(m_prior_check, type = "dens_overlay", ndraws = 100) +
  scale_x_log10()
```

This looks okay. We will not care about the astronomical values for now. The actual valuea re around 10^3 and some span from ^0 to ^5. It could be more limited but for weakly informative priors it should be ok.

```{r h1m3 torq sum model run}

if (file.exists(here(models, "h1.m3p_torq_cum.rds"))){
  
  h1.m3p_torq_cum <- readRDS(here(models, "h1.m3p_torq_cum.rds"))
  h1.m3p_torq_cum_R2 <- readRDS(here(models, "h1.m3p_torq_cum_R2.rds"))
  
} else {

  h1.m3p_torq_cum <- brm(
    arm_moment_sum_change_integral ~ 1 + correction_info + Modality_dummy + 
      BFI_extra + Familiarity + expressibility_z +  
      (1 + correction_info || pcn_ID) + (1 || SessionID) + 
      (1 + correction_info || concept) + (1 || TrialNumber_c),
    data = df_ges,
    family = lognormal(),
    prior = priors_h1m3p_torq_cum,
    iter = 2000,
    cores = 4
  )
  
  # Add criterions for later diagnostics
  h1.m3p_torq_cum <- add_criterion(h1.m3p_torq_cum, criterion = c("loo", "waic"))
  
  # Calculate also variance explained (R^2)
  h1.m3p_torq_cum_R2 <- bayes_R2(h1.m3p_torq_cum)
  
  # Save both as objects
  saveRDS(h1.m3p_torq_cum, here("models", "h1.m3p_torq_cum.rds"))
  saveRDS(h1.m3p_torq_cum_R2, here("models", "h1.m3p_torq_cum_R2.rds"))
  
}


# Summary
summary(h1.m3p_torq_cum)
# We still have Rhats>1 but we will deal with this later


```

These are the individual diagnostics

```{r h1m3p check}


plot(h1.m3p_torq_cum)
# all good

plot(conditional_effects(h1.m3p_torq_cum), points = TRUE)


pp_check(h1.m3p_torq_cum, type = "dens_overlay")


pp_check(h1.m3p_torq_cum, type = "error_scatter_avg")
# unchanged

h1.m3p_torq_cum_R2
# explained variance remains around 51%


```
Before doing the first round of diagnostics, let's also have a model with interactional terms

### Model 4 - Adding interactions

nteractions include:

-   CommAtt x Modality - Effort could increase differently across modalities, depending on whether concept is guessed on the first, second or third attempt. E.g., gesture might require more effort on initial attempt, but vocal require more effort in repeated attempt. This is interesting question and it would help disentangle the benfit of multimodality over unimodality (i.e., multimodal trials might be more effortful, but overal have less attempts). However, we already model effect of modality on effort, so maybe this is not top priority.

-   CommAtt x Expressibility - Higher expressibility should moderate the effect of repeated attempts, such that the increase in effort with each additional attempt is smaller (or bigger?) for more expressible concepts.This is not a priority for our analysis.

-   Modality × Expressibility_z - The influence of expressibility on effort could be modality-specific — perhaps effort increases less with expressibility in the combined modality. For our analysis, this is also not a priority (especially since expressibiliy has already modality encoded)

-   Familiarity x CommAtt - More familiar partners may guess faster (fewer attempts) and require less effort, but this effect could diminish over multiple attempts. For our analysis, this is not a priority,

-   Big5 × Modality or Big5 × CommAtt - More open/extraverted participants might maintain higher effort over attempts, or adjust more dynamically depending on the communicative channel. While not priority, it is an interesting question if we want to tap more into the inter-individual variability.

Let's try CommAtt x Modality and Big5 x CommAtt

```{r h1m4 torq cum run}

if (file.exists(here(models, "h1.m4_torq_cum.rds"))){
  
  h1.m4_torq_cum <- readRDS(here(models, "h1.m4_torq_cum.rds"))
  h1.m4_torq_cum_R2 <- readRDS(here(models, "h1.m4_torq_cum_R2.rds"))
  
} else {

  h1.m4_torq_cum <- brm(arm_moment_sum_change_integral ~ 1 + correction_info + Modality_dummy + correction_info:Modality_dummy + BFI_extra + correction_info:BFI_extra + Familiarity + expressibility_z + (1 + correction_info || pcn_ID) + (1 || SessionID) + (1 + correction_info || concept) + (1 || TrialNumber_c), 
                  data = df_ges,
                  family = lognormal(),
                  prior = priors_h1m3p_torq_cum, # we keep the priors from previous model
                  iter = 2000,
                  cores = 4)
  
  
  # Add criterions for later diagnostics
  h1.m4_torq_cum <- add_criterion(h1.m4_torq_cum, criterion = c("loo", "waic"))
  
  # Calculate also variance explained (R^2)
  h1.m4_torq_cum_R2 <- bayes_R2(h1.m4_torq_cum)
  
  # Save both as objects
  saveRDS(h1.m4_torq_cum, here(models, "h1.m4_torq_cum.rds"))
  saveRDS(h1.m4_torq_cum_R2, here(models, "h1.m4_torq_cum_R2.rds"))

}

# Summary
summary(h1.m4_torq_cum)
# We still have Rhats>1 but we will deal with this later




```

These are the individual diagnostics

```{r h1m4 torq cum check}


plot(h1.m4_torq_cum)
# all good

plot(conditional_effects(h1.m4_torq_cum), points = TRUE)


pp_check(h1.m4_torq_cum, type = "dens_overlay")


pp_check(h1.m4_torq_cum, type = "error_scatter_avg")
# unchanged

h1.m4_torq_cum_R2
# explained variance remains around 51%


```

### Diagnostics

We now have several models that we can test for predictive performance.

```{r h1 diagnostics}

model_list <- list(m0 = h1.m0_torq_cum, 
                   m1 = h1.m1_torq_cum, 
                   m2 = h1.m2_torq_cum, 
                   m3 = h1.m3_torq_cum, 
                   m3p = h1.m3p_torq_cum, 
                   m4 = h1.m4_torq_cum)
```


#### Rhat

Rhat tells us whether the model’s Markov chains have converged—values close to 1 mean the model has likely converged well.

```{r h1 rhat}

# Extract R-hat values for each model
rhat_list <- lapply(model_list, function(model) {
  rhat_values <- rhat(model)
  data.frame(model = deparse(substitute(model)), 
             max_rhat = max(rhat_values), 
             min_rhat = min(rhat_values))
})

# Combine and inspect
do.call(rbind, rhat_list)

```

First, second and 4th model seem the best.

#### ESS

Effective sample size tells how many independent samples the model has effectively drawn from the PD. Low ESS suggests autocorrelation (i.e., sample explores one part of posterior), while high ESS means good mix

```{r h1 ess}

# Extract n_eff values for each model
neff_ratio_list <- lapply(model_list, function(model) {
  neff_values <- neff_ratio(model)              # Here we calculate ratio (not the raw number of effective samples)
  data.frame(model = deparse(substitute(model)), 
             min_neff = min(neff_values), 
             max_neff = max(neff_values),
             mean_neff = mean(neff_values))
               
})

# Combine and inspect
do.call(rbind, neff_ratio_list)

```

The mean ratio is quite comparable but first two seem to be the best

```{r h1m0 torq cum ess}

effective_sample(h1.m0_torq_cum)

```
If we compare it with the m3p:

```{r h1m0 torq cum ess 2}

effective_sample(h1.m3p_torq_cum)

```
There is a big variability across the variables.

#### LOO & WAIC

Leave-One-Out (LOO) validation is a technique where, for each iteration, one data point is left out as the test set, and the model is trained on the remaining data points; this process is repeated for each data point, and the model’s overall performance is averaged over all iterations.

```{r h1 loo list}

l <- loo_compare(h1.m0_torq_cum, h1.m1_torq_cum, h1.m2_torq_cum, h1.m3_torq_cum, h1.m3p_torq_cum, h1.m4_torq_cum, criterion = "loo")

print(l, simplify = F)


```
::: callout-note
elpd_loo: This is the expected log pointwise predictive density for LOO. Higher values indicate a better fit to the data.

se_elpd_loo: The standard error of the elpd_loo, representing uncertainty in the model’s predictive fit according to LOO.

looic: The LOO Information Criterion, which is similar to waic but based on leave-one-out cross-validation. Lower values are better.

p_loo: The effective number of parameters according to LOO, indicating the model’s complexity.

se_p_loo: The standard error of p_loo, representing uncertainty around the effective number of parameters.
:::

Second model (with varying slopes and intercepts) has the best score as per LOO. 

```{r h1 waic I, message=FALSE, warning=FALSE}

w <- loo_compare(h1.m0_torq_cum, h1.m1_torq_cum, h1.m2_torq_cum, h1.m3_torq_cum, h1.m3p_torq_cum, h1.m4_torq_cum, criterion = "waic")

print(w, simplify = F)

# see Solomon Kurz
cbind(waic_diff = w[,1] * -2,
      se = w[,2] * 2)

```
::: callout-note
elpd_waic (expected log pointwise predictive density for WAIC): This represents the model's predictive fit to the data. Higher values indicate a better fit.

se_elpd_waic (standard error of elpd_waic): Measures uncertainty around the elpd_waic estimate.

waic: The Widely Applicable Information Criterion, a measure of model fit where lower values indicate a better fit.

se_waic (standard error of WAIC): Uncertainty around the WAIC estimate.

elpd_diff: The difference in the elpd_waic between the model in question and the baseline model (fit_eff_2, which has elpd_diff of 0). A negative value indicates that the model fits worse than fit_eff_2.

se_diff: The standard error of the elpd_diff, indicating how much uncertainty there is in the difference in predictive performance.

p_waic: The number of effective parameters in the model (related to model complexity). Lower values indicate simpler models, and higher values suggest more complexity.
:::

```{r h1 waic plot, echo=FALSE, message=FALSE, warning=FALSE}

w[, 7:8] %>% 
  data.frame() %>% 
  rownames_to_column("model_name") %>% 
  mutate(model_name = fct_reorder(model_name, waic, .desc = T)) %>% 
  
  ggplot(aes(x = waic, y = model_name, 
             xmin = waic - se_waic, 
             xmax = waic + se_waic)) +
  geom_pointrange(color = carto_pal(7, "BurgYl")[7], 
                  fill = carto_pal(7, "BurgYl")[5], shape = 21) +
  labs(title = "WAIC plot",
       x = NULL, y = NULL) +
  theme(axis.ticks.y = element_blank())

```

```{r h1 weights, message=FALSE, warning=FALSE}

model_weights(h1.m0_torq_cum, h1.m1_torq_cum, h1.m2_torq_cum, h1.m3_torq_cum, h1.m3p_torq_cum, h1.m4_torq_cum, weights = "waic") %>% 
  round(digits = 2)

```

Based on WAIC and LOO, model 2 (with varying slopes and intercept, with correlation but without interaction) seems to have the most predictive power. We can further finetune the priors as some of them are set to flat in model 2 which is not desirable.

Let's add the same priors we added to model 3


### Model 2.1 - Adding priors -RERUN (typo in priors)

```{r h1m2 priors}

prior_summary(h1.m2_torq_cum)
```

Now these are the same priors from model 3
```{r h1m2p torq cum new priors}


priors_h1m2p_torq_cum <- c(
  set_prior("normal(7, 1)", class = "Intercept"),
  set_prior("normal(0,0.4)", class = "b"),
  set_prior("normal(0,0.25)", class = "b", coef = "BFI_extra"),
  set_prior("normal(0,0.4)", class = "b", coef = "correction_info2M1"),
  set_prior("normal(0,0.4)", class = "b", coef = "correction_info3M2"),
  set_prior("normal(0,0.4)", class = "b", coef = "correction_info4M3"),
  set_prior("normal(0,0.35)", class = "b", coef = "expressibility_z"),
  set_prior("normal(0,0.25)", class = "b", coef = "Familiarity"),
  set_prior("normal(0,0.4)", class = "b", coef = "Modality_dummy1"), 
  set_prior("normal(0,0.15)", class = "sd"),
  set_prior("normal(0, 0.4)", class = "sigma"),
  # Random effect SDs: differentiated by grouping factor
  # Participants: largest expected source of variation
  set_prior("normal(0, 0.35)", class = "sd", group = "pcn_ID"),
  # Concepts: moderate
  set_prior("normal(0, 0.35)", class = "sd", group = "concept"),
  # Trial number: small drift
  set_prior("normal(0, 0.1)", class = "sd", group = "TrialNumber_c")
)

```

This will be then our model, but let's first just model the priors

```{r h1.m2 run prior only, eval=TRUE, message=FALSE, warning=FALSE}

h1.m2p_torq_cum_pps <- brm(arm_moment_sum_change_integral ~ 1 + correction_info + Modality_dummy + BFI_extra + Familiarity + expressibility_z +  (1 + correction_info | pcn_ID)  + (1 | SessionID) + (1 + correction_info | concept) + (1 | TrialNumber_c), 
                data = df_ges,
                prior = priors_h1m2p_torq_cum,
                sample_prior = "only",
                family = lognormal(),
                iter = 2000,
                cores = 4)

summary(h1.m2p_torq_cum_pps)

```

```{r h1.m2 prior summary}
prior_summary(h1.m2p_torq_cum_pps)
```


Now we do prior predictive check

```{r h1.m2 ppcheck}
pp_check(h1.m2p_torq_cum_pps, type = "dens_overlay", prefix = "ppd", ndraws = 100) +
  scale_x_log10()
```

Now with priors

```{r h1m2p torq cum model run}


if (file.exists(here(models, "h1.m2p_torq_cum.rds"))){
  
  h1.m2p_torq_cum <- readRDS(here(models, "h1.m2p_torq_cum.rds"))
  h1.m2p_torq_cum_R2 <- readRDS(here(models, "h1.m2p_torq_cum_R2.rds"))
  
} else {

  h1.m2p_torq_cum <- brm(arm_moment_sum_change_integral ~ 1 + correction_info + Modality_dummy + BFI_extra + Familiarity + expressibility_z +  (1 + correction_info | pcn_ID)  + (1 | SessionID) + (1 + correction_info | concept) + (1 | TrialNumber_c), 
                  data = df_ges,
                  prior = priors_h1m2p_torq_cum,
                  family = lognormal(),
                  iter = 2000,
                  cores = 4,
                  adapt_delta=0.95)
  
  
  # Add criterions for later diagnostics
  h1.m2p_torq_cum <- add_criterion(h1.m2p_torq_cum, criterion = c("loo", "waic"))
  
  # Calculate also variance explained (R^2)
  h1.m2p_torq_cum_R2 <- bayes_R2(h1.m2p_torq_cum)
  
  # Save both as objects
  saveRDS(h1.m2p_torq_cum, here(models, "h1.m2p_torq_cum"))
  saveRDS(h1.m2p_torq_cum_R2, here(models, "h1.m2p_torq_cum_R2.rds"))
  

}

# Summary
summary(h1.m2p_torq_cum)
# We still have Rhat 1.01 for sd of correction_info but the rest was solved. Some of the ESS are however still quite low


```

These are the individual diagnostics

```{r h1m2p torq cum check}


plot(h1.m2p_torq_cum)
# all good

plot(conditional_effects(h1.m2p_torq_cum), points = TRUE)


pp_check(h1.m2p_torq_cum, type = "dens_overlay", ndraws=100)


pp_check(h1.m2p_torq_cum, type = "error_scatter_avg")
# unchanged

h1.m2p_torq_cum_R2
# explained variance remains around 51%


```

### Model 5 (E) - Adding first effort

In the visualizations, it has become clear that it matters whether people put a lot of effort in c0 or not. We will therefore fit an additional, exploratory model to see whether this is reliable and how much does it add to the predictive & explanatory power

```{r h1m5 first effort merge}

# load in relevant data and merge with df_ges
c0_effort <- read_csv(paste0(featurefolder, "data_c0_effort.csv"))
df_ges_c0 <- full_join(df_ges, c0_effort, by = "TrialID")

# get rid of correction_info c0_only as this makes no sense here
df_ges_c0 <- df_ges_c0 |> filter(correction_info != 'c0_only')

# Propagate the c0 values to other rows (c1, c2)
c0_values <- df_ges_c0 |>
  filter(correction_info == "c0") |>
  select(pcn_ID, concept, effort_category_c0 = modality_specific_effort_category)

df_ges_c0 <- df_ges_c0 |>
  left_join(c0_values, by = c("pcn_ID", "concept")) |>
  mutate(
    modality_specific_effort_category = case_when(
      correction_info %in% c("c1", "c2") ~ effort_category_c0,
      TRUE ~ modality_specific_effort_category
    )
  ) |>
  select(-effort_category_c0)

# factorize relevant column
df_ges_c0$modality_specific_effort_category <- factor(df_ges_c0$modality_specific_effort_category, levels = c("Low Effort", "Medium Effort", "High Effort"))


# Now we also need to re-cc
df_ges_c0$correction_info <- factor(df_ges_c0$correction_info, 
                                   levels = c('c0', 'c1', 'c2'))

contrasts(df_ges_c0$correction_info) <- MASS::contr.sdif(3) 
#contrasts(final_data$modality) <- contr.sum(3)/2 # we will CC modality for each model separately

```

Note that we are adding effort category as an interaction term with correction_info, because adding it as a fixed factor makes no sense (unless a sannity check, we expect high effort category to have higher effort)


```{r h1m5 torq cum run}


priors_h1m5_torq_cum <- c(
  set_prior("normal(7, 1)", class = "Intercept"),
  set_prior("normal(0,0.4)", class = "b"),
  set_prior("normal(0,0.25)", class = "b", coef = "BFI_extra"),
  set_prior("normal(0,0.4)", class = "b", coef = "correction_info2M1"),
  set_prior("normal(0,0.4)", class = "b", coef = "correction_info3M2"),
  set_prior("normal(0,0.35)", class = "b", coef = "expressibility_z"),
  set_prior("normal(0,0.25)", class = "b", coef = "Familiarity"),
  set_prior("normal(0,0.4)", class = "b", coef = "Modality_dummy1"), 
  set_prior("normal(0,0.15)", class = "sd"),
  set_prior("normal(0, 0.4)", class = "sigma"),
  # Random effect SDs: differentiated by grouping factor
  # Participants: largest expected source of variation
  set_prior("normal(0, 0.35)", class = "sd", group = "pcn_ID"),
  # Concepts: moderate
  set_prior("normal(0, 0.35)", class = "sd", group = "concept"),
  # Trial number: small drift
  set_prior("normal(0, 0.1)", class = "sd", group = "TrialNumber_c")
)

if (file.exists(here(models, "h1.m5_torq_cum.rds"))){
  
  h1.m5_torq_cum <- readRDS(here(models, "h1.m5_torq_cum.rds"))
  h1.m5_torq_cum_R2 <- readRDS(here(models, "h1.m5_torq_cum_R2.rds"))
  
} else {
  
  h1.m5_torq_cum <- brm(arm_moment_sum_change_integral ~ 1 + correction_info * modality_specific_effort_category + Modality_dummy + BFI_extra + Familiarity + expressibility_z +  (1 + correction_info | pcn_ID)  + (1 | SessionID) + (1 + correction_info | concept) + (1 | TrialNumber_c), 
                data = df_ges_c0,
                prior = priors_h1m5_torq_cum,
                family = lognormal(),
                iter = 2000,
                cores = 4,
                adapt_delta=0.95)

  # Add criterions for later diagnostics
  h1.m5_torq_cum <- add_criterion(h1.m5_torq_cum, criterion = c("loo", "waic"))
  
  # Calculate also variance explained (R^2)
  h1.m5_torq_cum_R2 <- bayes_R2(h1.m5_torq_cum)
  
  # Save both as objects
  saveRDS(h1.m5_torq_cum, here(models, "h1.m5_torq_cum.rds"))
  saveRDS(h1.m5_torq_cum_R2, here(models, "h1.m5_torq_cum_R2.rds"))
  
}

# Summary
summary(h1.m5_torq_cum)
# Note that it massively improved the ESS

```

These are the individual diagnostics

```{r h1m5 torq cum check}


plot(h1.m5_torq_cum)
# all good

plot(conditional_effects(h1.m5_torq_cum), points = TRUE)


pp_check(h1.m5_torq_cum, type = "dens_overlay", ndraws=100)


pp_check(h1.m5_torq_cum, type = "error_scatter_avg")
# unchanged

h1.m5_torq_cum_R2
# explained variance improved slightly to 55%


```
### Model 6 (E) - Adding participant-level effort

Note: Maybe we can get this out from the random structure instead of fitting another model


## Var 2: Instant torque

```{r torq in hist}
hist(df_ges$arm_moment_sum_change_peak_mean)
```

### Model 0 - Intercept only

```{r h1m0 torq in run}

if (file.exists(here(models, "h1.m0_torq_in.rds"))){
  
  h1.m0_torq_in <- readRDS(here(models, "h1.m0_torq_in.rds"))
  h1.m0_torq_in_R2 <- readRDS(here(models, "h1.m0_torq_in_R2.rds"))
  
} else {
  
  h1.m0_torq_in <- brm(arm_moment_sum_change_peak_mean ~ 1 + correction_info, 
                data = df_ges,
                family = lognormal(),
                iter = 2000,
                cores = 4)
  
  # Add criterions for later diagnostics
  h1.m0_torq_in <- add_criterion(h1.m0_torq_in, criterion = c("loo", "waic"))
  
  # Calculate also variance explained (R^2)
  h1.m0_torq_in_R2 <- bayes_R2(h1.m0_torq_in)
  
  # Save both as objects
  saveRDS(h1.m0_torq_in, here(models, "h1.m0_torq_in.rds"))
  saveRDS(h1.m0_torq_in_R2, here(models, "h1.m0_torq_in_R2.rds"))

  
}

# Summary
summary(h1.m0_torq_in)

```
Check the diagnostics

```{r h1m0 torq in check}

plot(h1.m0_torq_in)
# all look ok

plot(conditional_effects(h1.m0_torq_in), points = TRUE)


pp_check(h1.m0_torq_in, type = "dens_overlay")
# Very good

pp_check(h1.m0_torq_in, type = "error_scatter_avg")
# The bigger the value, the bigger the residual error

h1.m0_torq_in_R2
# No captured variance 

```

### Model 1 - Simple reproduction of DAG 

First we want to do a simple model that reproduce our DAG.

Our main predictor is communicative attempt (CommAtt). To account for confounders - variables affecting both the predictor and the dependent variable - we need to adjust for them in the model by including them as covariates to isolate the effect of the predictor. Based on our DAG, confounders include:

1.  familiarity
2.  big5
3.  expressibility
4.  trial number
5.  modality
6.  concept
7.  participant

We include 1.-5. as fixed factors. For 6.-7., we include varying intercepts as we expect that each participant and concept may have they own baseline level of effort and thus allow for individual variation. Partial pooling is also beneficial in that extreme values (or categories will fewer datapoints) will be shrunk toward the overal average.

We will now not include PrevAn (previous answer) variable because we will need to do some further data-wrangling when building model for H2. That is mainly because PrevAn has some NA values, concretely for first attempts. Models would automatically exclude NA data, and we would therefore loose all effort data for attempt 1. For H1, however, we want to keep it there.

This is our first model without setting any priors, leaving them to default values


```{r h1.m1 model torq in, eval=TRUE, message=FALSE, warning=FALSE}

if (file.exists(here(models, "h1.m1_torq_in.rds"))){
  
  h1.m1_torq_in <- readRDS(here(models, "h1.m1_torq_in.rds"))
  h1.m1_torq_in_R2 <- readRDS(here(models, "h1.m1_torq_in_R2.rds"))
  
} else {
  
  h1.m1_torq_in <- brm(arm_moment_sum_change_peak_mean ~ 1 + correction_info + Familiarity + BFI_extra + expressibility_z + TrialNumber_c + Modality_dummy + (1 | pcn_ID) + (1 | SessionID) + (1 | concept),
                data = df_ges,
                family = lognormal(),
                iter = 2000,
                cores = 4)
  
  # Add criterions for later diagnostics
  h1.m1_torq_in <- add_criterion(h1.m1_torq_in, criterion = c("loo", "waic"))
  
  # Calculate also variance explained (R^2)
  h1.m1_torq_in_R2 <- bayes_R2(h1.m1_torq_in)
  
  # Save both as objects
  saveRDS(h1.m1_torq_in, here(models, "h1.m1_torq_in.rds"))
  saveRDS(h1.m1_torq_in_R2, here(models, "h1.m1_torq_in_R2.rds"))

  
}

# Summary
summary(h1.m1_torq_in)


```

Let's also check the visuals

```{r h1.m1 torq in check, echo=TRUE, message=FALSE, warning=FALSE}

plot(h1.m1_torq_in)
# b_Intercept looks a bit weird, but otherwise nothing alarming

plot(conditional_effects(h1.m1_torq_in), points = TRUE)


pp_check(h1.m1_torq_in, type = "dens_overlay")
# Looks ok

pp_check(h1.m1_torq_in, type = "error_scatter_avg")
# half blobby, half correlated, so still some room for improvement
# positive correlation means that errors increase with predicted values. So the model does perform well for some range, but becomes less reliable with increase in the predicted values
# also the blob is centered around 0 which is good
# it could be we are ignoring some interaction terms or non-linearity (which we know we kind of do). Transformation could also help (e.g., log). Of course, we are also still not specifying any priors so let's not yet make it a disaster

h1.m1_torq_in_R2
# explained variance around 50%

```

Overall, the perfomance is going in good direction

### Model 2 - Varying slopes and intercepts

```{r h1.m2 torq in run, eval=TRUE, message=FALSE, warning=FALSE}

if (file.exists(here(models, "h1.m2_torq_in.rds"))){
  
  h1.m2_torq_in <- readRDS(here(models, "h1.m2_torq_in.rds"))
  h1.m2_torq_in_R2 <- readRDS(here(models, "h1.m2_torq_in_R2.rds"))
  
} else {
  
  h1.m2_torq_in <- brm(arm_moment_sum_change_peak_mean ~ 1 + correction_info + Modality_dummy + BFI_extra + Familiarity + expressibility_z +  (1 + correction_info | pcn_ID)  + (1 | SessionID) + (1 + correction_info | concept) + (1 | TrialNumber_c), 
                data = df_ges,
                family = lognormal(),
                iter = 2000,
                cores = 4)
  
  # Add criterions for later diagnostics
  h1.m2_torq_in <- add_criterion(h1.m2_torq_in, criterion = c("loo", "waic"))
  
  # Calculate also variance explained (R^2)
  h1.m2_torq_in_R2 <- bayes_R2(h1.m2_torq_in)
  
  # Save both as objects
  saveRDS(h1.m2_torq_in, here(models, "h1.m2_torq_in.rds"))
  saveRDS(h1.m2_torq_in_R2, here(models, "h1.m2_torq_in_R2.rds"))
  
}

# Summary
summary(h1.m2_torq_in)

```


```{r h1.m2 torq in check, message=FALSE, warning=FALSE}

plot(h1.m2_torq_in)
# all look pretty fuzzy

plot(conditional_effects(h1.m2_torq_in), points = TRUE)

pp_check(h1.m2_torq_in, type = "dens_overlay")
# looks good

pp_check(h1.m2_torq_in, type = "error_scatter_avg")
# still positively correlated

h1.m2_torq_in_R2
# explained variance around 14%

```

### Model 3 - No correlation coefficient

```{r h1.m3 torq in run, eval=TRUE, message=FALSE, warning=FALSE}

if (file.exists(here(models, "h1.m3_torq_in.rds"))){
  
  h1.m3_torq_in <- readRDS(here(models, "h1.m3_torq_in.rds"))
  h1.m3_torq_in_R2 <- readRDS(here(models, "h1.m3_torq_in_R2.rds"))
  
} else {
  
  h1.m3_torq_in <- brm(arm_moment_sum_change_peak_mean ~ 1 + correction_info + Modality_dummy + BFI_extra + Familiarity + expressibility_z +  (1 + correction_info || pcn_ID) + (1 || SessionID) + (1 + correction_info || concept) + (1 || TrialNumber_c), 
                data = df_ges,
                family = lognormal(),
                iter = 2000,
                cores = 4)
  
  # Add criterions for later diagnostics
  h1.m3_torq_in <- add_criterion(h1.m3_torq_in, criterion = c("loo", "waic"))
  
  # Calculate also variance explained (R^2)
  h1.m3_torq_in_R2 <- bayes_R2(h1.m3_torq_in)
  
  # Save both as objects
  saveRDS(h1.m3_torq_in, here(models, "h1.m3_torq_in.rds"))
  saveRDS(h1.m3_torq_in_R2, here(models, "h1.m3_torq_in_R2.rds"))
  
}

# Summary
summary(h1.m3_torq_in)


```

```{r h1.m3 torq in check, message=FALSE, warning=FALSE}

plot(h1.m3_torq_in)
# all good

plot(conditional_effects(h1.m3_torq_in), points = TRUE)


pp_check(h1.m3_torq_in, type = "dens_overlay")


pp_check(h1.m3_torq_in, type = "error_scatter_avg")
# unchanged

h1.m3_torq_in_R2
# explained variance remains around 51%

```

### Model 3.1 - Adding priors

Now we will add some minimal priors, giving the model some reasonably informative, yet weak information.

These are the priors that were used for the last model

```{r h1m3 torq in priors}

prior_summary(h1.m3_torq_in)

```
We can see that for all betas, the priors are flat. We will put weakly informative priors to limit extreme values. Given that we have no strong opinion about the sigmas, we will update only those that were defaulted to flat priors, unless later it seems implausible.
We also set sigma because student-t(3,0,2.5) has heavy tails

We keep the priors as they were but change intercept because the values (and scale) are different

```{r h1m3 torq in set priors}

priors_h1m3p_torq_in <- c(
  set_prior("normal(-1, 0.5)", class = "Intercept"),
  set_prior("normal(0,0.4)", class = "b"),
  set_prior("normal(0,0.25)", class = "b", coef = "BFI_extra"),
  set_prior("normal(0,0.4)", class = "b", coef = "correction_info2M1"),
  set_prior("normal(0,0.4)", class = "b", coef = "correction_info3M2"),
  set_prior("normal(0,0.4)", class = "b", coef = "correction_info4M3"),
  set_prior("normal(0,0.35)", class = "b", coef = "expressibility_z"),
  set_prior("normal(0,0.25)", class = "b", coef = "Familiarity"),
  set_prior("normal(0,0.4)", class = "b", coef = "Modality_dummy1"), 
  set_prior("normal(0,0.15)", class = "sd"),
  set_prior("normal(0, 0.4)", class = "sigma"),
  # Random effect SDs: differentiated by grouping factor
  # Participants: largest expected source of variation
  set_prior("normal(0, 0.35)", class = "sd", group = "pcn_ID"),
  # Concepts: moderate
  set_prior("normal(0, 0.35)", class = "sd", group = "concept"),
  # Trial number: small drift
  set_prior("normal(0, 0.1)", class = "sd", group = "TrialNumber_c")
)


```

Now we do prior predictive check

```{r torq in prior check}

m_prior_check_in <- brm(
  arm_moment_sum_change_peak_mean ~ 1 + correction_info + Modality_dummy + 
    BFI_extra + Familiarity + expressibility_z +  
    (1 + correction_info || pcn_ID) + (1 || SessionID) + 
    (1 + correction_info || concept) + (1 || TrialNumber_c),
  data = df_ges,
  family = lognormal(),
  prior = priors_h1m3p_torq_in,
  sample_prior = "only",
  iter = 2000,
  cores = 4
)

pp_check(m_prior_check_in, ndraws = 1000, prefix = "ppd") +
  scale_x_log10()
```


This looks more or less like a reasonable range of values (even though 100 is slightly too extreme)
```{r h1m3p torq in run}

if (file.exists(here(models, "h1.m3p_torq_in.rds"))){
  
  h1.m3p_torq_in <- readRDS(here(models, "h1.m3p_torq_in.rds"))
  h1.m3p_torq_in_R2 <- readRDS(here(models, "h1.m3p_torq_in_R2.rds"))
  
} else {
  
  h1.m3p_torq_in <- brm(
  arm_moment_sum_change_peak_mean ~ 1 + correction_info + Modality_dummy + 
    BFI_extra + Familiarity + expressibility_z +  
    (1 + correction_info || pcn_ID) + (1 || SessionID) + 
    (1 + correction_info || concept) + (1 || TrialNumber_c),
  data = df_ges,
  family = lognormal(),
  prior = priors_h1m3p_torq_in,
  iter = 2000,
  cores = 4
)
  # Add criterions for later diagnostics
  h1.m3p_torq_in <- add_criterion(h1.m3p_torq_in, criterion = c("loo", "waic"))
  
  # Calculate also variance explained (R^2)
  h1.m3p_torq_in_R2 <- bayes_R2(h1.m3p_torq_in)
  
  # Save both as objects
  saveRDS(h1.m3p_torq_in, here(models, "h1.m3p_torq_in.rds"))
  saveRDS(h1.m3p_torq_in_R2, here(models, "h1.m3p_torq_in_R2.rds"))
  
}

# Summary
summary(h1.m3p_torq_in)


```

These are the individual diagnostics

```{r h1m3p torq in check}


plot(h1.m3p_torq_in)
# all good

plot(conditional_effects(h1.m3p_torq_in), points = TRUE)


pp_check(h1.m3p_torq_in, type = "dens_overlay", ndraws = 100)


pp_check(h1.m3p_torq_in, type = "error_scatter_avg")
# unchanged

h1.m3p_torq_in_R2
# explained variance remains around 14%


```
Before doing the first round of diagnostics, let's also have a model with interactional terms

### Model 4 - Adding interactions

nteractions include:

-   CommAtt x Modality - Effort could increase differently across modalities, depending on whether concept is guessed on the first, second or third attempt. E.g., gesture might require more effort on initial attempt, but vocal require more effort in repeated attempt. This is interesting question and it would help disentangle the benfit of multimodality over unimodality (i.e., multimodal trials might be more effortful, but overal have less attempts). However, we already model effect of modality on effort, so maybe this is not top priority.

-   CommAtt x Expressibility - Higher expressibility should moderate the effect of repeated attempts, such that the increase in effort with each additional attempt is smaller (or bigger?) for more expressible concepts.This is not a priority for our analysis. 

-   Modality × Expressibility_z - The influence of expressibility on effort could be modality-specific — perhaps effort increases less with expressibility in the combined modality. For our analysis, this is also not a priority (especially since expressibiliy has already modality encoded) 

-   Familiarity x CommAtt - More familiar partners may guess faster (fewer attempts) and require less effort, but this effect could diminish over multiple attempts. For our analysis, this is not a priority,

-   Big5 × Modality or Big5 × CommAtt - More open/extraverted participants might maintain higher effort over attempts, or adjust more dynamically depending on the communicative channel. While not priority, it is an interesting question if we want to tap more into the inter-individual variability.

Let's try CommAtt x Modality and Big5 x CommAtt

```{r h1m4 torq in run}

if (file.exists(here(models, "h1.m4_torq_in.rds"))){
  
  h1.m4_torq_in <- readRDS(here(models, "h1.m4_torq_in.rds"))
  h1.m4_torq_in_R2 <- readRDS(here(models, "h1.m4_torq_in_R2.rds"))
  
} else {
  
  h1.m4_torq_in <- brm(
  arm_moment_sum_change_peak_mean ~ 1 + correction_info + Modality_dummy + correction_info:Modality_dummy + BFI_extra + correction_info:BFI_extra + Familiarity + expressibility_z + (1 + correction_info || pcn_ID) + (1 || SessionID) + (1 + correction_info || concept) + (1 || TrialNumber_c),,
  data = df_ges,
  family = lognormal(),
  prior = priors_h1m3p_torq_in,
  iter = 2000,
  cores = 4
)
  # Add criterions for later diagnostics
  h1.m4_torq_in <- add_criterion(h1.m4_torq_in, criterion = c("loo", "waic"))
  
  # Calculate also variance explained (R^2)
  h1.m4_torq_in_R2 <- bayes_R2(h1.m4_torq_in)
  
  # Save both as objects
  saveRDS(h1.m4_torq_in, here(models, "h1.m4_torq_in.rds"))
  saveRDS(h1.m4_torq_in_R2, here(models, "h1.m4_torq_in_R2.rds"))
  
}

# Summary
summary(h1.m4_torq_in)


```

These are the individual diagnostics

```{r h1m4 torq in check}


plot(h1.m4_torq_in)
# all good

plot(conditional_effects(h1.m4_torq_in), points = TRUE)


pp_check(h1.m4_torq_in, type = "dens_overlay")


pp_check(h1.m4_torq_in, type = "error_scatter_avg")
# unchanged

h1.m4_torq_in_R2
# explained variance remains around 14%


```
Let's now also try the other: commatt & modality with expressibility 

### Model 5 - Interactions vol 2

```{r h1m5 torq in run}

if (file.exists(here(models, "h1.m5_torq_in.rds"))){
  
  h1.m5_torq_in <- readRDS(here(models, "h1.m5_torq_in.rds"))
  h1.m5_torq_in_R2 <- readRDS(here(models, "h1.m5_torq_in_R2.rds"))
  
} else {
  
  h1.m5_torq_in <- brm(
  arm_moment_sum_change_peak_mean ~ 1 + correction_info + Modality_dummy + BFI_extra  + Familiarity + expressibility_z + Modality_dummy:expressibility_z + correction_info:expressibility_z + (1 + correction_info || pcn_ID) + (1 || SessionID) + (1 + correction_info || concept) + (1 || TrialNumber_c),,
  data = df_ges,
  family = lognormal(),
  prior = priors_h1m3p_torq_in,
  iter = 2000,
  cores = 4
)
  # Add criterions for later diagnostics
  h1.m5_torq_in <- add_criterion(h1.m5_torq_in, criterion = c("loo", "waic"))
  
  # Calculate also variance explained (R^2)
  h1.m5_torq_in_R2 <- bayes_R2(h1.m5_torq_in)
  
  # Save both as objects
  saveRDS(h1.m5_torq_in, here(models, "h1.m5_torq_in.rds"))
  saveRDS(h1.m5_torq_in_R2, here(models, "h1.m5_torq_in_R2.rds"))
  
}

# Summary
summary(h1.m5_torq_in)


```
These are the individual diagnostics

```{r h1m5 torq in check}


plot(h1.m5_torq_in)
# all good

plot(conditional_effects(h1.m5_torq_in), points = TRUE)


pp_check(h1.m5_torq_in, type = "dens_overlay")


pp_check(h1.m5_torq_in, type = "error_scatter_avg")
# unchanged

h1.m5_torq_in_R2
# explained variance remains around 14%


```


### Diagnostics

We now have several models that we can test for predictive performance.

```{r h1 torq in model list}

model_list <- list(m0 = h1.m0_torq_in, 
                   m1 = h1.m1_torq_in, 
                   m2 = h1.m2_torq_in, 
                   m3 = h1.m3_torq_in, 
                   m3p = h1.m3p_torq_in, 
                   m4 = h1.m4_torq_in,
                   m5 = h1.m5_torq_in)
```


#### Rhat

Rhat tells us whether the model’s Markov chains have converged—values close to 1 mean the model has likely converged well.

```{r torq in rhat}

# Extract R-hat values for each model
rhat_list <- lapply(model_list, function(model) {
  rhat_values <- rhat(model)
  data.frame(model = deparse(substitute(model)), 
             max_rhat = max(rhat_values), 
             min_rhat = min(rhat_values))
})

# Combine and inspect
do.call(rbind, rhat_list)

```

Intercept-only model and full model with varying slopes and intercepts (without correlations, with priors) see, the best

#### ESS

Effective sample size tells how many independent samples the model has effectively drawn from the PD. Low ESS suggests autocorrelation (i.e., sample explores one part of posterior), while high ESS means good mix

```{r torq in ess}

# Extract n_eff values for each model
neff_ratio_list <- lapply(model_list, function(model) {
  neff_values <- neff_ratio(model)              # Here we calculate ratio (not the raw number of effective samples)
  data.frame(model = deparse(substitute(model)), 
             min_neff = min(neff_values), 
             max_neff = max(neff_values),
             mean_neff = mean(neff_values))
               
})

# Combine and inspect
do.call(rbind, neff_ratio_list)

```

The same models come out as the best

```{r torq in ess 1}

effective_sample(h1.m0_torq_in)

```
If we compare it with the m3p:

```{r torq in ess 2}

effective_sample(h1.m3p_torq_in)

```
From closer look they both seem comparable

#### LOO & WAIC

Leave-One-Out (LOO) validation is a technique where, for each iteration, one data point is left out as the test set, and the model is trained on the remaining data points; this process is repeated for each data point, and the model’s overall performance is averaged over all iterations.

```{r torq in loo}

l <- loo_compare(h1.m0_torq_in, h1.m1_torq_in, h1.m2_torq_in, h1.m3_torq_in, h1.m3p_torq_in, h1.m4_torq_in, h1.m5_torq_in, criterion = "loo")

print(l, simplify = F)


```
::: callout-note
elpd_loo: This is the expected log pointwise predictive density for LOO. Higher values indicate a better fit to the data.

se_elpd_loo: The standard error of the elpd_loo, representing uncertainty in the model’s predictive fit according to LOO.

looic: The LOO Information Criterion, which is similar to waic but based on leave-one-out cross-validation. Lower values are better.

p_loo: The effective number of parameters according to LOO, indicating the model’s complexity.

se_p_loo: The standard error of p_loo, representing uncertainty around the effective number of parameters.
:::

Second model (with varying slopes and intercepts) has the best score as per LOO. 

```{r h1 torq in waic I, message=FALSE, warning=FALSE}

w <- loo_compare(h1.m0_torq_in, h1.m1_torq_in, h1.m2_torq_in, h1.m3_torq_in, h1.m3p_torq_in, h1.m4_torq_in, h1.m5_torq_in, criterion = "waic")

print(w, simplify = F)

# see Solomon Kurz
cbind(waic_diff = w[,1] * -2,
      se = w[,2] * 2)

```
::: callout-note
elpd_waic (expected log pointwise predictive density for WAIC): This represents the model's predictive fit to the data. Higher values indicate a better fit.

se_elpd_waic (standard error of elpd_waic): Measures uncertainty around the elpd_waic estimate.

waic: The Widely Applicable Information Criterion, a measure of model fit where lower values indicate a better fit.

se_waic (standard error of WAIC): Uncertainty around the WAIC estimate.

elpd_diff: The difference in the elpd_waic between the model in question and the baseline model (fit_eff_2, which has elpd_diff of 0). A negative value indicates that the model fits worse than fit_eff_2.

se_diff: The standard error of the elpd_diff, indicating how much uncertainty there is in the difference in predictive performance.

p_waic: The number of effective parameters in the model (related to model complexity). Lower values indicate simpler models, and higher values suggest more complexity.
:::

```{r h1 torq in waic plot, echo=FALSE, message=FALSE, warning=FALSE}

w[, 7:8] %>% 
  data.frame() %>% 
  rownames_to_column("model_name") %>% 
  mutate(model_name = fct_reorder(model_name, waic, .desc = T)) %>% 
  
  ggplot(aes(x = waic, y = model_name, 
             xmin = waic - se_waic, 
             xmax = waic + se_waic)) +
  geom_pointrange(color = carto_pal(7, "BurgYl")[7], 
                  fill = carto_pal(7, "BurgYl")[5], shape = 21) +
  labs(title = "WAIC plot",
       x = NULL, y = NULL) +
  theme(axis.ticks.y = element_blank())

```

```{r h1 torq in weights, message=FALSE, warning=FALSE}

model_weights(h1.m0_torq_cum, h1.m1_torq_cum, h1.m2_torq_cum, h1.m3_torq_cum, h1.m3p_torq_cum, h1.m4_torq_cum, weights = "waic") %>% 
  round(digits = 2)

```

### Model 6 (E): Adding first effort

We will now do the same as in cumulative (model 5) but using the best performing model, i.e., Model 3.1

```{r h1m6 torq in prior}


priors_h1m6_torq_in <- c(
  set_prior("normal(-1, 0.5)", class = "Intercept"),
  set_prior("normal(0,0.4)", class = "b"),
  set_prior("normal(0,0.25)", class = "b", coef = "BFI_extra"),
  set_prior("normal(0,0.4)", class = "b", coef = "correction_info2M1"),
  set_prior("normal(0,0.4)", class = "b", coef = "correction_info3M2"),
  set_prior("normal(0,0.35)", class = "b", coef = "expressibility_z"),
  set_prior("normal(0,0.25)", class = "b", coef = "Familiarity"),
  set_prior("normal(0,0.4)", class = "b", coef = "Modality_dummy1"), 
  set_prior("normal(0,0.15)", class = "sd"),
  set_prior("normal(0, 0.4)", class = "sigma"),
  # Random effect SDs: differentiated by grouping factor
  # Participants: largest expected source of variation
  set_prior("normal(0, 0.35)", class = "sd", group = "pcn_ID"),
  # Concepts: moderate
  set_prior("normal(0, 0.35)", class = "sd", group = "concept"),
  # Trial number: small drift
  set_prior("normal(0, 0.1)", class = "sd", group = "TrialNumber_c")
)


```

```{r h1m6 torq in run}

if (file.exists(here(models, "h1.m6_torq_in.rds"))){
  
  h1.m6_torq_in <- readRDS(here(models, "h1.m6_torq_in.rds"))
  h1.m6_torq_in_R2 <- readRDS(here(models, "h1.m6_torq_in_R2.rds"))
  
} else {
  
  h1.m6_torq_in <- brm(
  arm_moment_sum_change_peak_mean ~ 1 + correction_info * modality_specific_effort_category + Modality_dummy + 
    BFI_extra + Familiarity + expressibility_z +  
    (1 + correction_info || pcn_ID) + (1 || SessionID) + 
    (1 + correction_info || concept) + (1 || TrialNumber_c),
  data = df_ges_c0,
  family = lognormal(),
  prior = priors_h1m6_torq_in,
  iter = 2000,
  cores = 4
)
  # Add criterions for later diagnostics
  h1.m6_torq_in <- add_criterion(h1.m6_torq_in, criterion = c("loo", "waic"))
  
  # Calculate also variance explained (R^2)
  h1.m6_torq_in_R2 <- bayes_R2(h1.m6_torq_in)
  
  # Save both as objects
  saveRDS(h1.m6_torq_in, here(models, "h1.m6_torq_in.rds"))
  saveRDS(h1.m6_torq_in_R2, here(models, "h1.m6_torq_in_R2.rds"))
  
}

# Summary
summary(h1.m6_torq_in)

```
These are the individual diagnostics

```{r h1m6 torq in check}


plot(h1.m6_torq_in)
# all good

plot(conditional_effects(h1.m6_torq_in), points = TRUE)


pp_check(h1.m6_torq_in, type = "dens_overlay", ndraws=100)


pp_check(h1.m6_torq_in, type = "error_scatter_avg")
# stil very much correlated

h1.m6_torq_in_R2
# explained variance improved slightly to 17%


```


# H2: Stating causal model

Now we can need to account also for our second hypothesis, namely

*H2: A higher degree of misunderstanding will require a performer to engage in more effortful correction.*

We operationalize degree of misunderstanding as a cosine similarity of performer's target and guesser's answer (PrevAn). The difficulty with our data structure is that PrevAn is a variable that has values only for second and third correction (i.e., there is no previous answer for the first performance). If we left it unchange, the model will eventually get rid of all NA values, which means we will loose the relationship between effort in second and first communicative attempt. To avoid this, we will create a new variable which we call *Effort Change*. We will simply calculate the change from effort in communicative attempt x to communicative attempt x+1. We will still loose communicative attempt 1, but the change that will be associated with CommAtt==2 already encapsulates the relationship towards this attempt.

We also need a new DAG. Essentially, what we said will influence CommAtt in H1, will now also influence PrevAn because they are tightly related. For instance, more extroverted people can be expected to be better guessers, therefore the similarity of the previous answer will be higher.

```{r dag h2, message=FALSE, warning=FALSE, echo=FALSE}

daggy_h2 <- dagitty('dag {
EffChange [outcome,pos="0,0"]
PrevAn [exposure,pos="-2,0"]
Big5 [pos="-1.5,1"]
CommAtt [pos="0,-1.5"]
Conc [pos="-2,-1.5"]
Expr [pos="-1.5,-1.5"]
Fam [pos="-0.5,1"]
Pcn [pos="-1,2"]
TrNum [pos="-2.5,-1.5"]
Mod [pos="-1,0.5"]

Big5 -> PrevAn     
Big5 -> EffChange
CommAtt -> EffChange
Conc -> Expr
Expr -> PrevAn
Expr -> EffChange
Fam -> PrevAn
Fam -> EffChange
Mod -> EffChange
Mod -> PrevAn
Pcn -> Big5
Pcn -> PrevAn 
Pcn -> EffChange
Pcn -> Fam
TrNum -> PrevAn
TrNum -> EffChange
Conc -> PrevAn
Conc -> EffChange
PrevAn -> EffChange
PrevAn -> CommAtt
}')

# Use ggdag for a cleaner visualization
d2 <- ggdag(daggy_h2, text = TRUE, use_labels = "name") +
  theme_dag() +
  ggtitle("Causal DAG for Effort Change") +
  theme(plot.title = element_text(hjust = 0.5))

d2

```

```{r dag h2 adjustset, message=FALSE, warning=FALSE, echo=TRUE}

dagitty::adjustmentSets(daggy_h2, exposure = "PrevAn", outcome = "EffChange")

```

The adjustment set is identical to the one of H1. Note that we are here omitting arrows going from all these variables to ComAtt. Since PrevAn affects CommAtt, and not the other way, we do not need to block this path to avoid confounds. However, if we want to asses direct effect of PrevAn on EffChange, we have to add CommAtt to the model as well.

# H2: Modeling

## Var 1: Cumulative torque

```{r h2 loading data, message=FALSE, warning=FALSE, include=FALSE}

# load our data back if we lost them
final_data <- data_feat

```

### Data wrangling I

```{r effort change ratio, message=FALSE, warning=FALSE}

# Calculate Effort Change (Difference)
final_data_2 <- final_data %>%
  group_by(pcn_ID, concept) %>%
  mutate(
    Effort_1 = arm_moment_sum_change_integral[correction_info == 'c0'][1],  # Effort for attempt 1
    Effort_2 = arm_moment_sum_change_integral[correction_info == 'c1'][1],  # Effort for attempt 2
    Effort_3 = arm_moment_sum_change_integral[correction_info == 'c2'][1],  # Effort for attempt 3
    Effort_Change_1_to_2 = case_when(
      correction_info == 'c1' & !is.na(Effort_1) ~ arm_moment_sum_change_integral - Effort_1,  
      # Change from 1st to 2nd attempt
      TRUE ~ NA_real_
    ),
    Effort_Change_2_to_3 = case_when(
      correction_info == 'c2' & !is.na(Effort_2) ~ arm_moment_sum_change_integral - Effort_2,  
      # Change from 2nd to 3rd attempt
      TRUE ~ NA_real_
    )
  ) %>%
  ungroup()

# Collide changes into a single column
final_data_2 <- final_data_2 %>%
  mutate(
    Effort_Change_arm_moment_integral = coalesce(Effort_Change_1_to_2, Effort_Change_2_to_3)
  ) 

# Remove unnecessary columns
final_data_2 <- subset(final_data_2, select = -c(Effort_1, Effort_2, Effort_3, Effort_Change_1_to_2, Effort_Change_2_to_3)) 

# View the result
head(final_data_2, n=15)

```

This is how effort change is distributed

```{r h2 hist, message=FALSE, warning=FALSE, echo=FALSE}

hist(final_data_2$Effort_Change_arm_moment_integral, breaks=100)

```

Convert columns to factors

```{r h2 factorizing, message=FALSE, warning=FALSE}

final_data <- final_data_2

# Create column TrialNumber by extracting third element from TrialID
final_data$TrialNumber <- sapply(strsplit(final_data$TrialID, "_"), function(x) x[3])
final_data$TrialNumber <- as.numeric(final_data$TrialNumber)  # Ensure TrialNumber is numeric

# Create sessionID by taking first element
final_data$SessionID <- sapply(strsplit(final_data$TrialID, "_"), function(x) x[1])
final_data$SessionID <- as.factor(final_data$SessionID)

# The rest as factor
final_data$correction_info <- as.factor(final_data$correction_info)
final_data$modality <- as.factor(final_data$modality)
final_data$pcn_ID <- as.factor(final_data$pcn_ID)
final_data$concept <- as.factor(final_data$concept)

```

Contrast-coding of categorical variables

```{r h2 contrast coding, message=FALSE, warning=FALSE}

# get rid of c0 and c0_only
#final_data <- final_data |> filter(correction_info != 'c0')
final_data <- final_data |> filter(correction_info != 'c0_only')
final_data <- final_data |> filter(correction_info != 'c0')

final_data$correction_info <- factor(final_data$correction_info, 
                                   levels = c('c1', 'c2'))

contrasts(final_data$correction_info) <- MASS::contr.sdif(2) 

```


Centering continuous variables


```{r h2 centering, message=FALSE, warning=FALSE}

final_data$TrialNumber_c <- final_data$TrialNumber - median(range(final_data$TrialNumber))
#final_data$Familiarity <- final_data$Familiarity - median(range(final_data$Familiarity))

# Z-score familiarity to be compatible with BFI
final_data$Familiarity <- scale(final_data$Familiarity, center = TRUE, scale = TRUE) |> as.numeric()

# FLAG: Not sure here about Familiarity. First, do I even treat it correctly (ie it is just sum of two questions), second: should it then be centered or z-scored


```

Note: BFI has been z-scored when processing metadata as this is the common procedure for calculating each trait's score

Z-scoring expressibility within a modality, and too cosine_similarity

```{r h2 zscoring, message=FALSE, warning=FALSE}
final_data <-
  final_data |>
  group_by(modality) |>
  mutate(expressibility_z = (expressibility - mean(expressibility))/ sd(final_data$expressibility, na.rm = T)) |>
  ungroup()
```


```{r h2 zscoring 2, message=FALSE, warning=FALSE}

range(final_data$answer_prev_dist)
hist(final_data$answer_prev_dist)

# Question: is z-scoring the correct way here?
final_data <-
  final_data |>
  mutate(answer_prev_dist_z = (answer_prev_dist - mean(answer_prev_dist))/ sd(final_data$answer_prev_dist, na.rm = T)) |>
  ungroup()

```

```{r}

# Get rid of vocal modality
df_ges <- final_data |> filter(modality != 'vocal')

# Dummy code modality
df_ges$Modality_dummy <- ifelse(df_ges$modality == "multimodal", 1, 0)
df_ges$Modality_dummy <- as.factor(df_ges$Modality_dummy)

```




<!-- z-scoring -->

Note: maybe we should center where we z-score to get better sense of the predictors (right now, for me it's still a bit uninterpretable)

<!-- z-scoring -->

### Model 0 - Intercept only

```{r}


if (file.exists(here(models, "h2.m0_torq_cum.rds"))){
  
  h2.m0_torq_cum <- readRDS(here(models, "h2.m0_torq_cum.rds"))
  h2.m0_torq_cum_R2 <- readRDS(here(models, "h2.m0_torq_cum_R2.rds"))
  
} else {
  
  h2.m0_torq_cum <- brm(Effort_Change_arm_moment_integral ~ 1 + answer_prev_dist_z, 
                data = df_ges,
                iter = 2000,
                cores = 4)
  
  # Add criterions for later diagnostics
  h2.m0_torq_cum <- add_criterion(h2.m0_torq_cum, criterion = c("loo", "waic"))
  
  # Calculate also variance explained (R^2)
  h2.m0_torq_cum_R2 <- bayes_R2(h2.m0_torq_cum)
  
  # Save both as objects
  saveRDS(h2.m0_torq_cum, here(models, "h2.m0_torq_cum.rds"))
  saveRDS(h2.m0_torq_cum_R2, here(models, "h2.m0_torq_cum_R2.rds"))

  
}

# Summary
summary(h2.m0_torq_cum)

```

Check the diagnostics

```{r}

plot(h2.m0_torq_cum)
# all look ok

plot(conditional_effects(h2.m0_torq_cum), points = TRUE)


pp_check(h2.m0_torq_cum, type = "dens_overlay")
# Not amazing fit 

pp_check(h2.m0_torq_cum, type = "error_scatter_avg")
# The bigger the value, the bigger the residual error

h2.m0_torq_cum_R2
# No captured variance 

```

### Model 1 - DAG

Similarly to H1, we will start with reproducing the DAG for H2

```{r h2.m1 run, eval=TRUE, message=FALSE, warning=FALSE}


if (file.exists(here(models, "h2.m1_torq_cum.rds"))){
  
  h2.m1_torq_cum <- readRDS(here(models, "h2.m1_torq_cum.rds"))
  h2.m1_torq_cum_R2 <- readRDS(here(models, "h2.m1_torq_cum_R2.rds"))
  
} else {
  
  h2.m1_torq_cum <- brm(Effort_Change_arm_moment_integral ~ 1 + answer_prev_dist_z + correction_info + Familiarity + BFI_extra + expressibility_z + TrialNumber_c + Modality_dummy + (1 | pcn_ID) + (1 | SessionID) + (1 | concept),
                data = df_ges,
                iter = 2000,
                cores = 4)

  
  # Add criterions for later diagnostics
  h2.m1_torq_cum <- add_criterion(h2.m1_torq_cum, criterion = c("loo", "waic"))
  
  # Calculate also variance explained (R^2)
  h2.m1_torq_cum_R2 <- bayes_R2(h2.m1_torq_cum)
  
  # Save both as objects
  saveRDS(h2.m1_torq_cum, here(models, "h2.m1_torq_cum.rds"))
  saveRDS(h2.m1_torq_cum_R2, here(models, "h2.m1_torq_cum_R2.rds"))

  
}

# Summary
summary(h2.m1_torq_cum)


```

```{r h2.m1 check, message=FALSE, warning=FALSE}

plot(h2.m1_torq_cum)
# all looks ok

plot(conditional_effects(h2.m1_torq_cum), points = TRUE)

pp_check(h2.m1_torq_cum, type = "dens_overlay")
# still of the distribution of the data

pp_check(h2.m1_torq_cum, type = "error_scatter_avg")
# so there seems to be quite high residual error

h2.m1_torq_cum_R2
# 2%

```
We should improve the family, because our predicted data are missing the shape of the real data's distribution. Given that the tails seem quite think, let's try student()

### Model 2 - Student-t distribution


```{r}


if (file.exists(here(models, "h2.m2_torq_cum.rds"))){
  
  h2.m2_torq_cum <- readRDS(here(models, "h2.m2_torq_cum.rds"))
  h2.m2_torq_cum_R2 <- readRDS(here(models, "h2.m2_torq_cum_R2.rds"))
  
} else {
  
  h2.m2_torq_cum <- brm(Effort_Change_arm_moment_integral ~ 1 + answer_prev_dist_z + correction_info + Familiarity + BFI_extra + expressibility_z + TrialNumber_c + Modality_dummy + (1 | pcn_ID) + (1 | SessionID) + (1 | concept),
                data = df_ges,
                family = student(),
                iter = 2000,
                cores = 4)

  
  # Add criterions for later diagnostics
  h2.m2_torq_cum <- add_criterion(h2.m2_torq_cum, criterion = c("loo", "waic"))
  
  # Calculate also variance explained (R^2)
  h2.m2_torq_cum_R2 <- bayes_R2(h2.m2_torq_cum)
  
  # Save both as objects
  saveRDS(h2.m2_torq_cum, here(models, "h2.m2_torq_cum.rds"))
  saveRDS(h2.m2_torq_cum_R2, here(models, "h2.m2_torq_cum_R2.rds"))

  
}

# Summary
summary(h2.m2_torq_cum)


```
Diagnostics

```{r h2.m2 torq cum check, message=FALSE, warning=FALSE}

plot(h2.m2_torq_cum)
# all looks ok

plot(conditional_effects(h2.m2_torq_cum), points = TRUE)

pp_check(h2.m2_torq_cum, type = "dens_overlay", ndraws=100)
# this is much better, even though now we have some crazy values (esp on the right), which we can probably regularize by priors

pp_check(h2.m2_torq_cum, type = "error_scatter_avg")
# so there seems to be quite high residual error

h2.m2_torq_cum_R2
# drop to 0

```

### Model 3 - Varying slopes and intercepts

We are now going to procede with the same steps as in models for H1. 

Do we have enough data points?

```{r}
df_ges |>
  group_by(pcn_ID) |>
  summarise(n = n()) |>
  summary()
# Need reasonable spread of answer_prev_dist_c within each participant
# to estimate a slope reliably

df_ges |>
  group_by(pcn_ID) |>
  summarise(sd_dist = sd(answer_prev_dist_z)) |>
  summary()
# If sd_dist is near 0 for many participants, the slope is not identifiable


```
Let's add random slopes. 

```{r}

if (file.exists(here(models, "h2.m3_torq_cum.rds"))){
  
  h2.m3_torq_cum <- readRDS(here(models, "h2.m3_torq_cum.rds"))
  h2.m3_torq_cum_R2 <- readRDS(here(models, "h2.m3_torq_cum_R2.rds"))
  
} else {
  
  h2.m3_torq_cum <- brm(Effort_Change_arm_moment_integral ~ 1 + answer_prev_dist_z + correction_info + Familiarity + BFI_extra + expressibility_z + Modality_dummy + (1 + answer_prev_dist_z | pcn_ID)  + (1 | SessionID) + (1 + answer_prev_dist_z | concept) + (1 | TrialNumber_c), 
                data = df_ges,
                family = student(),
                iter = 2000,
                cores = 4)
  
  # Add criterions for later diagnostics
  h2.m3_torq_cum <- add_criterion(h2.m3_torq_cum, criterion = c("loo", "waic"))
  
  # Calculate also variance explained (R^2)
  h2.m3_torq_cum_R2 <- bayes_R2(h2.m3_torq_cum)
  
  # Save both as objects
  saveRDS(h2.m3_torq_cum, here(models, "h2.m3_torq_cum.rds"))
  saveRDS(h2.m3_torq_cum_R2, here(models, "h2.m3_torq_cum_R2.rds"))
  
}

# Summary
summary(h2.m3_torq_cum)



```

Diagnostics

```{r h2.m1 torq cum check, message=FALSE, warning=FALSE}

plot(h2.m3_torq_cum)
# all looks ok

plot(conditional_effects(h2.m3_torq_cum), points = TRUE)

pp_check(h2.m3_torq_cum, type = "dens_overlay", ndraws=100)
# this is much better, even though now we have some crazy values (esp on the right), which we can probably regularize by priors

pp_check(h2.m3_torq_cum, type = "error_scatter_avg")
#  still very high

h2.m3_torq_cum_R2
# 1%

```

### Model 3.1 - Adding priors


```{r}

prior_summary(h2.m3_torq_cum)

```
```{r}

priors_h2m3p_torq_in <- c(
  set_prior("normal(0, 1000)", class = "Intercept"),
  set_prior("normal(0,200)", class = "b", coef = "answer_prev_dist_z"),
  set_prior("normal(0,200)", class = "b", coef = "BFI_extra"),
  set_prior("normal(0,200)", class = "b", coef = "correction_info2M1"),
  set_prior("normal(0,200)", class = "b", coef = "expressibility_z"),
  set_prior("normal(0,200)", class = "b", coef = "Familiarity"),
  set_prior("normal(0,200)", class = "b", coef = "Modality_dummy1"), 
  set_prior("normal(0,200)", class = "sd"),
  set_prior("normal(0,1000)", class = "sigma"),
  # Random effect SDs: differentiated by grouping factor
  # Participants: largest expected source of variation
  set_prior("normal(0, 500)", class = "sd", group = "pcn_ID"),
  # Concepts: moderate
  set_prior("normal(0, 500)", class = "sd", group = "concept"),
  # Trial number: small drift
  set_prior("normal(0, 200)", class = "sd", group = "TrialNumber_c")
)

```

Now we can fit the priors

```{r}


# h2.m3p_torq_cum_pps <- brm(Effort_Change_arm_moment_integral ~ 1 + answer_prev_dist_z + correction_info + Familiarity + BFI_extra + expressibility_z + Modality_dummy + (1 + answer_prev_dist_z | pcn_ID)  + (1 | SessionID) + (1 + answer_prev_dist_z | concept) + (1 | TrialNumber_c), 
#                 data = df_ges,
#                 family = student(),
#                 iter = 2000,
#                 prior = priors_h2m3p_torq_in,
#                 sample_prior = "only",
#                 cores = 4)
# 
# pp_check(h2.m3p_torq_cum_pps, prefix="ppd", ndraws=100)

```

This looks good, now we can fit model
```{r}

if (file.exists(here(models, "h2.m3p_torq_cum.rds"))){
  
  h2.m3p_torq_cum <- readRDS(here(models, "h2.m3p_torq_cum.rds"))
  h2.m3p_torq_cum_R2 <- readRDS(here(models, "h2.m3p_torq_cum_R2.rds"))
  
} else {
  
  h2.m3p_torq_cum <- brm(Effort_Change_arm_moment_integral ~ 1 + answer_prev_dist_z + correction_info + Familiarity + BFI_extra + expressibility_z + Modality_dummy + (1 + answer_prev_dist_z | pcn_ID)  + (1 | SessionID) + (1 + answer_prev_dist_z | concept) + (1 | TrialNumber_c), 
                data = df_ges,
                family = student(),
                iter = 2000,
                prior = priors_h2m3p_torq_in,
                cores = 4)
  
  # Add criterions for later diagnostics
  h2.m3p_torq_cum <- add_criterion(h2.m3p_torq_cum, criterion = c("loo", "waic"))
  
  # Calculate also variance explained (R^2)
  h2.m3p_torq_cum_R2 <- bayes_R2(h2.m3p_torq_cum)
  
  # Save both as objects
  saveRDS(h2.m3p_torq_cum, here(models, "h2.m3p_torq_cum.rds"))
  saveRDS(h2.m3p_torq_cum_R2, here(models, "h2.m3p_torq_cum_R2.rds"))
  
}

# Summary
summary(h2.m3p_torq_cum)



```
Diagnostics

```{r}

plot(h2.m3p_torq_cum)
# all looks ok

plot(conditional_effects(h2.m3p_torq_cum), points = TRUE)

pp_check(h2.m3p_torq_cum, type = "dens_overlay", ndraws=100)
# we still have some crazy values :3

pp_check(h2.m3p_torq_cum, type = "error_scatter_avg")
#  not really better here

h2.m3p_torq_cum_R2
# 1%

```

### Model 4 - Adding interactions

Last thing we do with linear models is that we add interactions. As mentioned in the preregistrated script, we can consider:

-   PrevAn x Modality - similarity affects the change in effort differently (e.g., vocal might still require more effort)

-   PrevAn x Expressibility - similarity in relation to effort might matter only for highly expressible concepts, and low expressible concepts are difficult to express, so also difficult to exaggerate


```{r}

if (file.exists(here(models, "h2.m4_torq_cum.rds"))){
  
  h2.m4_torq_cum <- readRDS(here(models, "h2.m4_torq_cum.rds"))
  h2.m4_torq_cum_R2 <- readRDS(here(models, "h2.m4_torq_cum_R2.rds"))
  
} else {
  
  h2.m4_torq_cum <- brm(Effort_Change_arm_moment_integral ~ 1 + answer_prev_dist_z + correction_info + Familiarity + BFI_extra + expressibility_z + answer_prev_dist_z:expressibility_z +  Modality_dummy + answer_prev_dist_z:Modality_dummy + (1 + answer_prev_dist_z | pcn_ID)  + (1 | SessionID) + (1 + answer_prev_dist_z | concept) + (1 | TrialNumber_c), 
                data = df_ges,
                family = student(),
                iter = 2000,
                prior = priors_h2m3p_torq_in,
                cores = 4)
  
  # Add criterions for later diagnostics
  h2.m4_torq_cum <- add_criterion(h2.m4_torq_cum, criterion = c("loo", "waic"))
  
  # Calculate also variance explained (R^2)
  h2.m4_torq_cum_R2 <- bayes_R2(h2.m4_torq_cum)
  
  # Save both as objects
  saveRDS(h2.m4_torq_cum, here(models, "h2.m4_torq_cum.rds"))
  saveRDS(h2.m4_torq_cum_R2, here(models, "h2.m4_torq_cum_R2.rds"))
  
}

# Summary
summary(h2.m4_torq_cum)



```
Diagnostics

```{r}

plot(h2.m4_torq_cum)
# all looks ok

plot(conditional_effects(h2.m4_torq_cum), points = TRUE)

pp_check(h2.m4_torq_cum, type = "dens_overlay", ndraws=100)
# we still have some crazy values :3

pp_check(h2.m4_torq_cum, type = "error_scatter_avg")
#  not really better here

h2.m4_torq_cum_R2
# 1%

```

### Splines 1 - fitting b-splines

In the previous model for H1, we were able to spot possible non-linear trends by contrast coding of the communicative attempts whereby each level was defined relative to the two remaining. Now, we are dealing with situation where both outcome variable and the main predictor are continuous variable. Also here we might expect nonlinear patterns in their relationship. For instance, it might be the case that for answers that are completely off, people do not feel motivated to put in more effort, and some kind of sensible 'threshold' needs to be reached for the effort to increase to resolve misunderstanding, and then again decrease when the answer is very similar.

We will use a combination of B-splines and Bayesian GAMs to prepare some models and try them fit to the synthetic data. 

```{r h2 plot nonlinear, echo=FALSE, message=FALSE, warning=FALSE}

ggplot(df_ges, aes(x = answer_prev_dist, y = Effort_Change_arm_moment_integral)) +
  geom_point(alpha = 0.6, color = "blue") +  # Scatter points
  geom_smooth(method = "loess", color = "red", se = FALSE) +  # loess for potential nonlinearity
  labs(x = "Previous Answer Similarity (PrevAn)", 
       y = "Effort (Eff)", 
       title = "Relationship between Effort and Previous Answer Similarity") +
  theme_minimal()

```

B-splines allow to build up wiggly functions from more simple components which are called 'basis functions' (B). They divide a full range of predictor into parts and assign a parameter to each part. The parameters are gradually turned on and off in a way that makes their sum into a curve. Unlike polynomial regression, b-splines do not transform predictor, but they invent a series of new synthetic predictor variables. Each of them then exists only to gradually turn a specific parameter on and off within a specific range of the real predictor variable. Each of these synthetic variables is called a basis function B. See more in Richard McElreath's Statistical Rethinking [@mcelreath18].

We are using the [code](https://bookdown.org/content/4857/geocentric-models.html) from Solomon Kurz's adaptation of this book.

```{r h2 datawrangle, message=FALSE, warning=FALSE}

# Get rid of NAs in the predictor
d <-
  df_ges %>% 
  drop_na(answer_prev_dist)

# And convert all that is necessary to factor/numerical
#d$CommAtt <- as.factor(d$CommAtt)
#d$Modality <- as.factor(d$Modality)
#d$Participant <- as.factor(d$Participant)
#d$Concept <- as.factor(d$Concept)
#d$TrialNumber <- as.numeric(d$TrialNumber) 

```

Here we can see summary of each predictor

```{r h2 datawrangle 2, message=FALSE, warning=FALSE}

d %>% 
  select_if(is.numeric) %>%  # Select only numeric columns
  pivot_longer(cols = everything(), names_to = "key", values_to = "value") %>%
  group_by(key) %>%
  summarise(
    mean = mean(value, na.rm = TRUE),
    sd   = sd(value, na.rm = TRUE),
    ll   = quantile(value, probs = 0.055, na.rm = TRUE),
    ul   = quantile(value, probs = 0.945, na.rm = TRUE)
  ) %>%
  mutate(across(where(is.double), round, digits = 2))

head(d, n=10)

```

Now we need to specify knots that function as pivots for number of different basis functions. The B variable then tells you which knot you are close to.

```{r h2 knots, message=FALSE, warning=FALSE}

num_knots <- 7
knot_list <- quantile(d$answer_prev_dist_z, probs = seq(from = 0, to = 1, length.out = num_knots))
knot_list

```

Now we can see how we chopped the data by the knots

```{r h2 knot plot, message=FALSE, warning=FALSE, echo=FALSE}

d %>% 
  ggplot(aes(x = answer_prev_dist_z, y = Effort_Change_arm_moment_integral)) +
  geom_vline(xintercept = knot_list, color = "white", alpha = 1/2) +
  geom_point(color = "#ffb7c5", alpha = 1/2) +
  theme_bw() +
  theme(panel.background = element_rect(fill = "#4f455c"),
        panel.grid = element_blank())

```

Now we need to specify the polynomial degree which determines how parameters interact to produce the spline.

For degree 1, two basis functions combine at each point. For degree 2, three functions combine at each point. For degree 3, four combine.

```{r b-spline, message=FALSE, warning=FALSE}

B <- bs(d$answer_prev_dist_z,
        knots = knot_list[-c(1, num_knots)], 
        degree = 3, # cubic spline
        intercept = TRUE)

```

This is how cubic spline with 7 knots look like

```{r b-spline plot, message=FALSE, warning=FALSE, echo=FALSE}

# wrangle a bit
b <-
  B %>% 
  data.frame() %>% 
  set_names(str_c(0, 1:4), 5:9) %>%  
  bind_cols(select(d, answer_prev_dist_z)) %>% 
  pivot_longer(-answer_prev_dist_z,
               names_to = "bias_function",
               values_to = "bias")

# plot
b %>% 
  ggplot(aes(x = answer_prev_dist_z, y = bias, group = bias_function)) +
  geom_vline(xintercept = knot_list, color = "white", alpha = 1/2) +
  geom_line(color = "#ffb7c5", alpha = 1/2, linewidth = 1.5) +
  ylab("bias value") +
  theme_bw() +
  theme(panel.background = element_rect(fill = "#4f455c"),
        panel.grid = element_blank())

```

We will now add this B matrix into the data to be able to further use it in the models

```{r b-spline 2, message=FALSE, warning=FALSE}

d2 <-
  d %>% 
  mutate(B = B) 

# take a look at the structure of `d3
d2 %>% glimpse()

```

The B matrix is now a matrix column which contains the same number of rows as the others, but has also 9 columns within that columns. Each of them correspond to one synthetic B variable.

We can now use this matrix to fit our model

```{r h2.s1 run, eval=TRUE, message=FALSE, warning=FALSE}

if (file.exists(here(models, "h2.s1_torq_cum.rds"))){
  
  h2.s1_torq_cum <- readRDS(here(models, "h2.s1_torq_cum.rds"))
  h2.s1_torq_cum_R2 <- readRDS(here(models, "h2.s1_torq_cum_R2.rds"))
  
} else {

  h2.s1_torq_cum <- 
    brm(data = d2,
        family = gaussian,
        Effort_Change_arm_moment_integral ~ 1 + B,
        prior = c(prior(normal(100, 10), class = Intercept),
                  prior(normal(0, 10), class = b),
                  prior(exponential(1), class = sigma)),
        iter = 4000, warmup = 2000, chains = 4, cores = 4,
        seed = 4)

  
  # Add criterions for later diagnostics
  h2.s1_torq_cum <- add_criterion(h2.s1_torq_cum, criterion = c("loo", "waic"))
  
  # Calculate also variance explained (R^2)
  h2.s1_torq_cum_R2 <- bayes_R2(h2.s1_torq_cum)
  
  # Save both as objects
  saveRDS(h2.s1_torq_cum, here(models, "h2.s1_torq_cum.rds"))
  saveRDS(h2.s1_torq_cum_R2, here(models, "h2.s1_torq_cum_R2.rds"))
  
}

print(h2.s1_torq_cum)

```

It's a bit difficult to see what is going on so let's just plot it

```{r h2.s1 plot, message=FALSE, warning=FALSE, echo=FALSE}

post <- as_draws_df(h2.s1_torq_cum)

post %>% 
  select(b_B1:b_B9) %>% 
  set_names(c(str_c(0, 1:4), 5:9)) %>% 
  pivot_longer(everything(), names_to = "bias_function") %>% 
  group_by(bias_function) %>% 
  summarise(weight = mean(value)) %>% 
  full_join(b, by = "bias_function") %>% 
  
  # plot
  ggplot(aes(x = answer_prev_dist_z, y = bias * weight, group = bias_function)) +
  geom_vline(xintercept = knot_list, color = "white", alpha = 1/2) +
  geom_line(color = "#ffb7c5", alpha = 1/2, linewidth = 1.5) +
  theme_bw() +
  theme(panel.background = element_rect(fill = "#4f455c"),
        panel.grid = element_blank()) 

```

Now with the predictor

```{r h2.s1 plot2, message=FALSE, warning=FALSE, echo=FALSE}

f <- fitted(h2.s1_torq_cum)

f %>% 
  data.frame() %>% 
  bind_cols(d2) %>% 
  
  ggplot(aes(x = answer_prev_dist_z, y = Effort_Change_arm_moment_integral, ymin = Q2.5, ymax = Q97.5)) + 
  geom_vline(xintercept = knot_list, color = "white", alpha = 1/2) +
  geom_hline(yintercept = fixef(h2.s1_torq_cum)[1, 1], color = "white", linetype = 2) +
  geom_point(color = "#ffb7c5", alpha = 1/2) +
  geom_ribbon(fill = "white", alpha = 2/3) +
  labs(x = "Previous Answer",
       y = "Effort Change") +
  theme_bw() +
  theme(panel.background = element_rect(fill = "#4f455c"),
        panel.grid = element_blank())

```

Now let's check some diagnostics

```{r h2.s1 check, message=FALSE, warning=FALSE}

plot(h2.s1_torq_cum)
# looks good

pp_check(h2.s1_torq_cum, type = "dens_overlay")
# because we used only main predictor, we can see quite a bad result of the posterior predictive distribution, but we will work on that

pp_check(h2.s1_torq_cum, type = "error_scatter_avg")
# high residual error for both low and high values

h2.s1_torq_cum_R2
# explained variance essentialy 0

```

### Splines 2 - Bayesian GAMs

Now we will use a different workflow, using smooth functions with `brms` package. Brms allow for non-linear models by borrowing functions from `mgcv` package [@wood17]

This is how priors should look like for one of these functions

```{r h2.s2 priors, message=FALSE, warning=FALSE}

get_prior(data = d2,
          family = gaussian,
          Effort_Change_arm_moment_integral ~ 1 + s(answer_prev_dist_z))
```

At this point, I will just keep priors at default

```{r h2.s2 run, eval=TRUE, message=FALSE, warning=FALSE}

if (file.exists(here(models, "h2.s2_torq_cum.rds"))){
  
  h2.s2_torq_cum <- readRDS(here(models, "h2.s2_torq_cum.rds"))
  h2.s2_torq_cum_R2 <- readRDS(here(models, "h2.s2_torq_cum_R2.rds"))
  
} else {
  
  h2.s2_torq_cum <-
    brm(data = d2,
        family = student(),
        Effort_Change_arm_moment_integral ~ 1 + s(answer_prev_dist_z, bs = "bs", k = 19),
        iter = 2000, 
        warmup = 1000, 
        chains = 4, 
        cores = 4,
        seed = 4,
        control = list(adapt_delta = .99)
        )

  # Add criterions for later diagnostics
  h2.s2_torq_cum <- add_criterion(h2.s2_torq_cum, criterion = c("loo", "waic"))
  
  # Calculate also variance explained (R^2)
  h2.s2_torq_cum_R2 <- bayes_R2(h2.s2_torq_cum)
  
  # Save both as objects
  saveRDS(h2.s2_torq_cum, here(models, "h2.s2_torq_cum.rds"))
  saveRDS(h2.s2_torq_cum_R2, here(models, "h2.s2_torq_cum_R2.rds"))

  }

# Summary
summary(h2.s2_torq_cum)

```

Now we can proceed as we usually do

```{r h2.s2 check, message=FALSE, warning=FALSE}

plot(h2.s2_torq_cum)
# looks good

plot(conditional_effects(h2.s2_torq_cum), points = TRUE)
# again, we see similar (non-linear) trend 

pp_check(h2.s2_torq_cum, type = "dens_overlay")
# looks good

pp_check(h2.s2_torq_cum, type = "error_scatter_avg")
# still weird

h2.s2_torq_cum_R2
# explained variance 2%

```

### Splines 3 - GAMs with f+r effects

Now let's use the same model, but adding our other predictors

```{r h2.s3 run, eval=TRUE, message=FALSE, warning=FALSE}

if (file.exists(here(models, "h2.s3_torq_cum.rds"))){
  
  h2.s3_torq_cum <- readRDS(here(models, "h2.s3_torq_cum.rds"))
  h2.s3_torq_cum_R2 <- readRDS(here(models, "h2.s3_torq_cum_R2.rds"))
  
} else {
  
  h2.s3_torq_cum <- 
    brm(
      data = d2,
      family = student(),
      Effort_Change_arm_moment_integral ~ 1 +
        s(answer_prev_dist_z, bs = "bs", k = 19) +  # Smooth for Previous Answer similarity
        correction_info + Familiarity + BFI_extra + expressibility_z + Modality_dummy + 
        (1 + answer_prev_dist_z | pcn_ID)  + (1 | SessionID) + (1 + answer_prev_dist_z | concept) + (1 | TrialNumber_c),
      iter = 2000, 
      warmup = 1000, 
      chains = 4, 
      cores = 4,
      seed = 4,
      control = list(adapt_delta = .99)
    )

  
  # Add criterions for later diagnostics
  h2.s3_torq_cum <- add_criterion(h2.s3_torq_cum, criterion = c("loo", "waic"))
  
  # Calculate also variance explained (R^2)
  h2.s3_torq_cum_R2 <- bayes_R2(h2.s3_torq_cum)
  
  # Save both as objects
  saveRDS(h2.s3_torq_cum, here(models, "h2.s3_torq_cum.rds"))
  saveRDS(h2.s3_torq_cum_R2, here(models, "h2.s3_torq_cum_R2.rds"))

}

summary(h2.s3_torq_cum)


```

```{r h2.s3 check, message=FALSE, warning=FALSE}


plot(h2.s3_torq_cum)
# looks ok

plot(conditional_effects(h2.s3_torq_cum), points = TRUE)
# now the ffect of PrevAn looks a bit different because we are back to using z-scored version

pp_check(h2.s3_torq_cum, type = "dens_overlay")
# looks good

pp_check(h2.s3_torq_cum, type = "error_scatter_avg")
# still not what we would like

h2.s3_torq_cum_R2
# 1 %


```

### Model 5 - Previous effort as a predictor instead of change

```{r}

df_ges <- df_ges |>
  arrange(pcn_ID, concept, TrialNumber) |>
  group_by(pcn_ID, concept) |>
  mutate(
    effort_previous_moment_integral = lag(arm_moment_sum_change_integral, n = 1)
  ) |>
  ungroup()

# Test

df_ges |>
  filter(pcn_ID == df_ges$pcn_ID[1], concept == df_ges$concept[1]) |>
  select(pcn_ID, concept, TrialNumber, correction_info, 
         arm_moment_sum_change_integral, effort_previous_moment_integral)


# Re-cc
df_ges <- df_ges |> filter(correction_info != 'c0')
df_ges$correction_info <- factor(df_ges$correction_info, 
                                   levels = c('c1', 'c2'))

contrasts(df_ges$correction_info) <- MASS::contr.sdif(2) 

# z-score again 

df_ges <-
  df_ges |>
  mutate(answer_prev_dist_z = (answer_prev_dist - mean(answer_prev_dist))/ sd(df_ges$answer_prev_dist, na.rm = T)) |>
  ungroup()

# center previous effort
df_ges <- df_ges |>
  mutate(
    effort_previous_moment_integral_c = effort_previous_moment_integral - mean(effort_previous_moment_integral, na.rm = TRUE)
  )

```


We will use the priors from H1, and add priors for cosine_similarity and previous_effort


```{r}


priors_h2m5_torq_cum <- c(
  set_prior("normal(5, 2)", class = "Intercept"),
  set_prior("normal(0,0.5)", class = "b", coef = "answer_prev_dist_z"), # new
  set_prior("normal(0,0.25)", class = "b", coef = "BFI_extra"),
  set_prior("normal(0,0.5)", class = "b", coef = "correction_info2M1"),
  set_prior("normal(0,0.5)", class = "b", coef = "effort_previous_moment_integral_c"), # new
  set_prior("normal(0,0.35)", class = "b", coef = "expressibility_z"),
  set_prior("normal(0,0.25)", class = "b", coef = "Familiarity"),
  set_prior("normal(0,0.5)", class = "b", coef = "Modality_dummy1"),
  set_prior("normal(0,0.15)", class = "sd"),
  set_prior("normal(0, 0.5)", class = "sigma"),
  # Random effect SDs: differentiated by grouping factor
  # Participants: largest expected source of variation
  set_prior("normal(0, 0.35)", class = "sd", group = "pcn_ID"),
  # Concepts: moderate
  set_prior("normal(0, 0.35)", class = "sd", group = "concept"),
  # Trial number: small drift
  set_prior("normal(0, 0.1)", class = "sd", group = "TrialNumber_c")
)
  
```

Now we can fit the model with effort as IV, and previous_effort as predictor in interaction with the answer. 

```{r}


if (file.exists(here(models, "h2.m5_torq_cum.rds"))){
  
  h2.m5_torq_cum <- readRDS(here(models, "h2.m5_torq_cum.rds"))
  h2.m5_torq_cum_R2 <- readRDS(here(models, "h2.m5_torq_cum_R2.rds"))
  
} else {
  
  h2.m5_torq_cum <- brm(arm_moment_sum_change_integral ~ 1 + answer_prev_dist_z * effort_previous_moment_integral_c + correction_info + Familiarity + BFI_extra + expressibility_z + Modality_dummy + (1 + answer_prev_dist_z | pcn_ID)  + (1 | SessionID) + (1 + answer_prev_dist_z | concept) + (1 | TrialNumber_c),
                data = df_ges,
                family = lognormal(),
                prior = priors_h2m5_torq_cum,
                iter = 2000,
                cores = 4)

  
  # Add criterions for later diagnostics
  h2.m5_torq_cum <- add_criterion(h2.m5_torq_cum, criterion = c("loo", "waic"))
  
  # Calculate also variance explained (R^2)
  h2.m5_torq_cum_R2 <- bayes_R2(h2.m5_torq_cum)
  
  # Save both as objects
  saveRDS(h2.m5_torq_cum, here(models, "h2.m5_torq_cum.rds"))
  saveRDS(h2.m5_torq_cum_R2, here(models, "h2.m5_torq_cum_R2.rds"))

  
}

# Summary
summary(h2.m5_torq_cum)


```

Note: the Rhats are worrying...

Diagnostics

```{r}


plot(h2.m5_torq_cum)
# looks ok

plot(conditional_effects(h2.m5_torq_cum), points = TRUE)

pp_check(h2.m5_torq_cum, type = "dens_overlay")
# looks good

pp_check(h2.m5_torq_cum, type = "error_scatter_avg")
# This looks much better!

h2.m5_torq_cum_R2
# 59 % !!!


```
Does it make sense to fit non-linear in this way?

```{r}

ggplot(df_ges, aes(x = answer_prev_dist, y = arm_moment_sum_change_integral)) +
  geom_point(alpha = 0.6, color = "blue") +  # Scatr points
  geom_smooth(method = "loess", color = "red", se = FALSE) +  # loess for potential nonlinearity
  labs(x = "Previous Answer Similarity (PrevAn)", 
       y = "Effort (Eff)", 
       title = "Relationship between Effort and Previous Answer Similarity") +
  theme_minimal()

```
Not really


### Diagnostics I

Before we proceed further, let's do first round of diagnostics

```{r h2 diagnostics, include=FALSE, message=FALSE, warning=FALSE}

model_list <- list(h2.m0_torq_cum, h2.m1_torq_cum, h2.m2_torq_cum, h2.m3_torq_cum, h2.m3p_torq_cum, h2.m4_torq_cum, h2.s1_torq_cum, h2.s2_torq_cum, h2.s3_torq_cum, h2.m5_torq_cum)

```

#### Rhat

```{r h2 rhat, message=FALSE, warning=FALSE}

# Extract R-hat values for each model
rhat_list <- lapply(model_list, function(model) {
  rhat_values <- rhat(model)
  data.frame(model = deparse(substitute(model)), 
             max_rhat = max(rhat_values), 
             min_rhat = min(rhat_values))
})

# Combine and inspect
do.call(rbind, rhat_list)

```

All Rhat values look good except the last one

### ESS

Effective sample size tells how many independent samples the model has effectively drawn from the PD. Low ESS suggests autocorrelation (i.e., sample explores one part of posterior), while high ESS means good mix

```{r h2 ess, message=FALSE, warning=FALSE}

# Extract n_eff values for each model
neff_ratio_list <- lapply(model_list, function(model) {
  neff_values <- neff_ratio(model)              # Here we calculate ratio (not the raw number of effective samples)
  data.frame(model = deparse(substitute(model)), 
             min_neff = min(neff_values), 
             max_neff = max(neff_values),
             mean_neff = mean(neff_values))
               
})

# Combine and inspect
do.call(rbind, neff_ratio_list)

```

They all look good expect the last one again

### LOO & WAIC

```{r h2 loo, message=FALSE, warning=FALSE}

l <- loo_compare(h2.m0_torq_cum, h2.m1_torq_cum, h2.m2_torq_cum, h2.m3_torq_cum, h2.m3p_torq_cum, h2.m4_torq_cum, h2.s1_torq_cum, h2.s2_torq_cum, h2.s3_torq_cum, h2.m5_torq_cum, criterion = "loo")

print(l, simplify = F)

```
Assessed by LOO, here we see that the last one (with previous_effort as FF) is actually the best.

```{r h2 waic, message=FALSE, warning=FALSE}

w <- loo_compare(h2.m0_torq_cum, h2.m1_torq_cum, h2.m2_torq_cum, h2.m3_torq_cum, h2.m3p_torq_cum, h2.m4_torq_cum, h2.s1_torq_cum, h2.s2_torq_cum, h2.s3_torq_cum, h2.m5_torq_cum, criterion = "waic")

print(w, simplify = F)

# see Solomon Kurz
cbind(waic_diff = w[,1] * -2,
      se = w[,2] * 2)

```

Plot the comparison

```{r h2 waic plot, message=FALSE, warning=FALSE, echo=FALSE}

w[, 7:8] %>% 
  data.frame() %>% 
  rownames_to_column("model_name") %>% 
  mutate(model_name = fct_reorder(model_name, waic, .desc = T)) %>% 
  
  ggplot(aes(x = waic, y = model_name, 
             xmin = waic - se_waic, 
             xmax = waic + se_waic)) +
  geom_pointrange(color = carto_pal(7, "BurgYl")[7], 
                  fill = carto_pal(7, "BurgYl")[5], shape = 21) +
  labs(title = "WAIC plot",
       x = NULL, y = NULL) +
  theme(axis.ticks.y = element_blank())

```

Here we see identical results

```{r h2 weights, message=FALSE, warning=FALSE}

model_weights(h2.m0_torq_cum, h2.m1_torq_cum, h2.m2_torq_cum, h2.m3_torq_cum, h2.m3p_torq_cum, h2.m4_torq_cum, h2.s1_torq_cum, h2.s2_torq_cum, h2.s3_torq_cum, h2.m5_torq_cum, weights = "waic") %>% 
  round(digits = 2)

```
## Var 2: Instant torque

### Data wrangling

```{r}

df_ges <- df_ges |>
  arrange(pcn_ID, concept, TrialNumber) |>
  group_by(pcn_ID, concept) |>
  mutate(
    effort_previous_moment_peak_mean = lag(arm_moment_sum_change_peak_mean, n = 1)
  ) |>
  ungroup()

# Re-cc
df_ges <- df_ges |> filter(correction_info != 'c0')
df_ges <- df_ges |> filter(correction_info != 'c0_only')
df_ges$correction_info <- factor(df_ges$correction_info, 
                                   levels = c('c1', 'c2'))

contrasts(df_ges$correction_info) <- MASS::contr.sdif(2) 

# z-score again 
df_ges <-
  df_ges |>
  mutate(answer_prev_dist_z = (answer_prev_dist - mean(answer_prev_dist))/ sd(df_ges$answer_prev_dist, na.rm = T)) |>
  ungroup()

# center previous effort
df_ges <- df_ges |>
  mutate(
    effort_previous_moment_peak_mean_c = effort_previous_moment_peak_mean - mean(effort_previous_moment_peak_mean, na.rm = TRUE)
  )

```

### Model 5 - Previous effort

We will use the priors from H1, and add priors for cosine_similarity and previous_effort

```{r}


priors_h2m5_torq_in <- c(
  set_prior("normal(-1, 0.5)", class = "Intercept"),
  set_prior("normal(0,0.5)", class = "b", coef = "answer_prev_dist_z"), # new
  set_prior("normal(0,0.25)", class = "b", coef = "BFI_extra"),
  set_prior("normal(0,0.5)", class = "b", coef = "correction_info2M1"),
  set_prior("normal(0,0.5)", class = "b", coef = "effort_previous_moment_peak_mean_c"), # new
  set_prior("normal(0,0.35)", class = "b", coef = "expressibility_z"),
  set_prior("normal(0,0.25)", class = "b", coef = "Familiarity"),
  set_prior("normal(0,0.5)", class = "b", coef = "Modality_dummy1"),
  set_prior("normal(0,0.15)", class = "sd"),
  set_prior("normal(0, 0.5)", class = "sigma"),
  # Random effect SDs: differentiated by grouping factor
  # Participants: largest expected source of variation
  set_prior("normal(0, 0.35)", class = "sd", group = "pcn_ID"),
  # Concepts: moderate
  set_prior("normal(0, 0.35)", class = "sd", group = "concept"),
  # Trial number: small drift
  set_prior("normal(0, 0.1)", class = "sd", group = "TrialNumber_c")
)

  
```

Now we can fit the model with effort as IV, and previous_effort as predictor in interaction with the answer. 

```{r}


if (file.exists(here(models, "h2.m5_torq_in.rds"))){
  
  h2.m5_torq_in <- readRDS(here(models, "h2.m5_torq_in.rds"))
  h2.m5_torq_in_R2 <- readRDS(here(models, "h2.m5_torq_in_R2.rds"))
  
} else {
  
  h2.m5_torq_in <- brm(arm_moment_sum_change_peak_mean ~ 1 + answer_prev_dist_z * effort_previous_moment_peak_mean_c + correction_info + Familiarity + BFI_extra + expressibility_z + Modality_dummy + (1 + answer_prev_dist_z | pcn_ID)  + (1 | SessionID) + (1 + answer_prev_dist_z | concept) + (1 | TrialNumber_c),
                data = df_ges,
                family = lognormal(),
                prior = priors_h2m5_torq_in,
                iter = 2000,
                cores = 4)

  
  # Add criterions for later diagnostics
  h2.m5_torq_in <- add_criterion(h2.m5_torq_in, criterion = c("loo", "waic"))
  
  # Calculate also variance explained (R^2)
  h2.m5_torq_in_R2 <- bayes_R2(h2.m5_torq_in)
  
  # Save both as objects
  saveRDS(h2.m5_torq_in, here(models, "h2.m5_torq_in.rds"))
  saveRDS(h2.m5_torq_in_R2, here(models, "h2.m5_torq_in_R2.rds"))

  
}

# Summary
summary(h2.m5_torq_in)


```
Interestingly here we have no problems with Rhat (perhaps because the values are smaller?)

Diagnostics

```{r}


plot(h2.m5_torq_in)
# looks ok

plot(conditional_effects(h2.m5_torq_in), points = TRUE)

pp_check(h2.m5_torq_in, type = "dens_overlay", ndraws=100)
# looks good

pp_check(h2.m5_torq_in, type = "error_scatter_avg")
# still very positively correlated

h2.m5_torq_in_R2
# 11 % This is much less than the cumulative model - why?


```
Does it make sense to fit non-linear in this way?

```{r}

ggplot(df_ges, aes(x = answer_prev_dist_z, y = arm_moment_sum_change_peak_mean)) +
  geom_point(alpha = 0.6, color = "blue") +  # Scatr points
  geom_smooth(method = "loess", color = "red", se = FALSE) +  # loess for potential nonlinearity
  labs(x = "Previous Answer Similarity (PrevAn)", 
       y = "Effort (Eff)", 
       title = "Relationship between Effort and Previous Answer Similarity") +
  theme_minimal()

```
Not really. But we can try interactions as previously


+ answer_prev_dist_z:expressibility_z +  Modality_dummy + answer_prev_dist_z:Modality_dummy 

### Model 6 - Adding interactions

```{r}


if (file.exists(here(models, "h2.m6_torq_in.rds"))){
  
  h2.m6_torq_in <- readRDS(here(models, "h2.m6_torq_in.rds"))
  h2.m6_torq_in_R2 <- readRDS(here(models, "h2.m6_torq_in_R2.rds"))
  
} else {
  
  h2.m6_torq_in <- brm(arm_moment_sum_change_peak_mean ~ 1 + answer_prev_dist_z * effort_previous_moment_peak_mean_c + correction_info + Familiarity + BFI_extra + expressibility_z + answer_prev_dist_z:expressibility_z +  Modality_dummy + answer_prev_dist_z:Modality_dummy + (1 + answer_prev_dist_z | pcn_ID)  + (1 | SessionID) + (1 + answer_prev_dist_z | concept) + (1 | TrialNumber_c),
                data = df_ges,
                family = lognormal(),
                prior = priors_h2m5_torq_in,
                iter = 2000,
                cores = 4)

  
  # Add criterions for later diagnostics
  h2.m6_torq_in <- add_criterion(h2.m6_torq_in, criterion = c("loo", "waic"))
  
  # Calculate also variance explained (R^2)
  h2.m6_torq_in_R2 <- bayes_R2(h2.m6_torq_in)
  
  # Save both as objects
  saveRDS(h2.m6_torq_in, here(models, "h2.m6_torq_in.rds"))
  saveRDS(h2.m6_torq_in_R2, here(models, "h2.m6_torq_in_R2.rds"))

  
}

# Summary
summary(h2.m6_torq_in)


```

Diagnostics

```{r}


plot(h2.m6_torq_in)
# looks ok

plot(conditional_effects(h2.m6_torq_in), points = TRUE)

pp_check(h2.m6_torq_in, type = "dens_overlay", ndraws=100)
# looks good

pp_check(h2.m6_torq_in, type = "error_scatter_avg")
# still very positively correlated

h2.m6_torq_in_R2
# 11 % 


```


# Session info

Here is session information for reproducibility: 

```{r session info print, echo=FALSE, message=FALSE, warning=FALSE}

sessionInfo()

```


# Questions for Daniela

- c0_only, should it stay or should it go when not necessarily of interest (or merge with c0?)
- participant variability, how to explore from model
- concept variability, how to explore from model
- should we worry about low R2?


